---
title: "Primary Analysis"
author: "Jonathan Hwang"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Summary

This notebook cleans the mortality data from the 1630-1631 Venetian Plague, calculates the Shannon entropy and negentropy of the quantile function and the PDF, and uses bootstrapping to find 95% confidence intervals.

# Initialization

Initialize renv and load packages.

```{r setup, include=TRUE, warning=FALSE,message=FALSE}
renv::init()

library(tidyverse)
library(rootSolve)
library(data.table)
library(msos)
library(entropy)
library(simboot)
library(MCPAN)
library(boot)
library(gridExtra)
library(cowplot)
library(ggpubr)
library(progress)
library(renv)
library(gridExtra)
library(Hmisc)
library(lubridate)
library(tmap)
library(ggplotify)
library(sf)
sf_use_s2(FALSE)

# Clear environment
rm(list = ls())

knitr::opts_chunk$set(echo = TRUE)
```

Import all of the [csv files](https://github.com/ggrrll/Venice-plague-epidemic-paper) containing mortality data sourced from *Lazzari, G., Colavizza, G., Bortoluzzi, F., Drago, D., Erboso, A., Zugno, F., ... & Salathé, M. (2020). A digital reconstruction of the 1630–1631 large plague outbreak in Venice. Scientific reports, 10(1), 1-7*.

```{r import mortality csvs}
# Mortality data for all parishes from January 1, 1629 to December 31, 1631.
all <- read.csv("ALL_utf.csv")
city_peaks <- read.csv("city_cases_baseline_removed.csv")
filled <- read.csv("filledna_full_ts.csv")

# Geographical data for locations of Venetian parishes. 
# Note that geo_df doesn't have S.Giacomo.dell'Orio and Angelo.Rafael is blank.
geo_df <- read.csv("geo_df.csv") 

# Venetian census data from 1696 containing parish populations
population <- read.csv("ParishPopulations.csv")

# 1586 census data containing which parishes belong to which sestieri 
sestiere_pre <- read.csv("sestiereDataPre.csv")

# 1696 census data containing which parishes belong to which sestieri 
sestiere_post <- read.csv("SestiereDataPost.csv")
```

Import csv files containing population data from Venetian censuses in 1509, 1551, 1581, 1586, and 1696 for each parish.

```{r import_population_csvs}
population_pre <- read.csv("ParishPopulationsPre.csv")
population_post <- read.csv("ParishPopulationsPost.csv")
```

# Data Cleaning

Now, we perform preliminary cleaning.

```{r first_pass_cleaning}
# Fix the Dates on all
all$Date <- sub("2029", "1629", all$Date)
all$Date <- sub("2030", "1630", all$Date)
all$Date <- sub("2031", "1631", all$Date)
all$Date <- as.Date(all$Date, format = "%m/%d/%Y")

# Fix artifacts for column names
colnames(all) <- sub("^S\\.", "S", colnames(all))

# Convert mortality data in all into integer types
for(column in 1:ncol(all)){
  if(column == 1){
    next
  }
  if(!(is.integer(all[,column]))){
    all[,column] = as.integer(all[,column])
  }
}

# Fix the Dates on filled
colnames(filled)[1] <- "Date"
filled$Date <- as.Date(filled$Date, format = "%Y-%m-%d")

# Fix the parish name artifacts in filled (import_csv replaces spaces with periods)
colnames(filled) <- sub("^S\\.", "S", colnames(filled))

# Convert mortality data in filled into integer types
for(column in 1:ncol(filled)){
  if(column == 1){
    next
  }
  if(!(is.integer(filled[, column]))){
    filled[, column] = as.integer(filled[,column])
  }
}

# Fix the Dates on city_peaks
colnames(city_peaks)[1] <- "Date"
city_peaks$Date <- sub(" 00:00:00", "", city_peaks$Date)
city_peaks$Date <- as.Date(city_peaks$Date, format = "%Y-%m-%d")

# Fix geo_df
colnames(geo_df)[1] <- "Parish"
geo_df$Parish <- gsub("[[:space:]]", ".", geo_df$Parish)
geo_df$Parish <- sub("^S\\.", "S", geo_df$Parish)

# Get rid of Angelo.Rafael because it has no lat/long data
# geo_df <- geo_df[!grepl("Angelo.Rafael",geo_df$Parish),]

# Fix population
population$Parish <- gsub("[[:space:]]", ".", population$Parish)
population$Parish <- sub("^S\\.", "S", population$Parish)

# What are the missing cities in geo_df?
setdiff(geo_df$Parish, colnames(filled))

# What are the missing cities in population?
filled <- filled[, -match(c("S.Marcilian"), colnames(filled))] # Marcilian only exists in the population_pre dataset.

# Remove columns (parishes) in filled that exclusively contain NAs
all_nas <- c()
for(col in 1:ncol(filled)){
  if(sum(is.na(filled[,col])) == length(filled[,col])){
    all_nas <- c(all_nas, col)
  }
}
if(length(all_nas) > 0){
  filled <- filled[, -all_nas]
}

# Remove parishes in filled that aren't in the population dataset
missing_parishes <- c()
for(p in 1:length(colnames(filled))){
  if(!(colnames(filled)[p] %in% population$Parish) & colnames(filled)[p] != "Date"){
    missing_parishes <- c(missing_parishes, p)
  }
}
if(length(missing_parishes) > 0){
  filled <- filled[,-missing_parishes]
}
```

Clean the population datasets.

```{r population_cleaning}
colnames(population_pre) <- c("Parish", "Population_1509", "Population_1551", 
                              "Population_1581", "Population_1586")

# Remove artifacts
population_pre$Parish <- gsub("[[:space:]]", ".", population_pre$Parish)
population_pre$Parish <- sub("^S\\.", "S", population_pre$Parish)

population_post$Parish <- gsub("[[:space:]]", ".", population_post$Parish)
population_post$Parish <- sub("^S\\.", "S", population_post$Parish)

# Identify and remove the parishes that are not contained in both the 
# 1586 census and the 1696 census
remove_vec <- c()
for(i in 1:length(population_pre$Parish)){
  if(!(population_pre$Parish[i] %in% population_post$Parish)){
    remove_vec <- c(remove_vec,i)
  }
}

if(length(remove_vec) > 0){
  population_pre <- population_pre[-remove_vec,]
}

remove_vec <- c()
for(i in 1:length(population_post$Parish)){
  if(!(population_post$Parish[i] %in% population_pre$Parish)){
    remove_vec <- c(remove_vec, i)
  }
}

if(length(remove_vec) > 0){
  population_post <- population_post[-remove_vec,]
}

# Establish a consistent ordering of the parishes (alphabetical)
population_post <- population_post %>% arrange(Parish)
population_pre <- population_pre %>% arrange(Parish)

# Remove the 1509 census data because the vast majority of values are NAs
population_pre <- population_pre[,-2]

# population_pre dataset is in thousands of people, so multiply by a thousand
# to get the population

for(j in 1:ncol(population_pre)){
  if(j == 1){
    next
  }
  for(i in 1:nrow(population_pre)){
    population_pre[i,j] <- format(as.numeric(population_pre[i,j]), nsmall = 3)
    
    # Some parishes already have their population multiplied by a thousand, and
    # those parishes will have population numbers with 7 or more characters
    # because we include 3 decimal places in our call of the format function.
    # Thus, we only scale the numbers that have less than 7 characters.
    
    if(nchar(population_pre[i,j]) < 7){ 
      population_pre[i,j] <- as.numeric(population_pre[i,j]) * 1000
    } else {
      population_pre[i,j] <- as.numeric(population_pre[i,j])
    }
  }
  population_pre[,j] <- as.numeric(population_pre[,j])
}
population_post$Population <- as.numeric(population_post$Population) * 1000

# Merge population_pre and population_post into one dataframe, population_df
population_df <- as.data.frame(cbind(population_pre, population_post[,-1]))
colnames(population_df) <- c("Parish", "Population_1551", "Population_1581", 
                             "Population_1586", "Population_1696")

# Plot the population data over time
year_vec <- c(c(1551, 1581, 1586, 1696))

t_pop_df <- transpose(population_df)
rownames(t_pop_df) <- colnames(population_df)
t_pop_df <- t_pop_df[-1,]
colnames(t_pop_df) <- population_df$Parish
t_pop_df <- cbind(year_vec, t_pop_df)
t_pop_df_plot <- t_pop_df %>% 
  gather(key = "Parish_var", value = "value", -year_vec)
t_pop_df_plot$value <- as.numeric(t_pop_df_plot$value)

t_pop_df_plot %>%
  ggplot(aes(x = year_vec, y = value))+
  geom_line(aes(color = Parish_var), show.legend = FALSE, linewidth = 0.5)+
  labs(x = "Year", y = "Population", 
       title = "Population over Time for each Parish")+
  scale_y_log10()+
  theme_minimal()

# ggsave("PopulationsOverTimeNoLegendYLog10.png", width = 1920, height = 1080, units = "px")
```

Plot the 1696 population data vs. the 1586 population data.

```{r population_plot}
population_df %>%
  ggplot(aes(x = Population_1586, y = Population_1696))+
  geom_point(size = 0.5)+
  geom_smooth(method = lm, se = FALSE, linewidth = 0.5)+
  theme_bw()+
  theme(text = element_text(family = ""), panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  labs(x = "Population in 1586", y = "Population in 1696", title = "Population in 1696 vs. Population in 1586")

# Save as Supplementary Figure 1 (now 6)
# ggsave("SuppFig1.jpg", width = 1920, height = 1080, units = "px")
```

Run a linear regression on the Venetian census data from 1696 vs. 1586 to determine the validity of using linear interpolation to estimate 1629 parish populations.

```{r linear_regression}
cor(population_df$Population_1586, population_df$Population_1696) #0.95

plot(x = population_df$Population_1586, y = population_df$Population_1696)

abline(mod <- lm(population_df$Population_1696 ~ population_df$Population_1586)) 
# y = 1696, x = 1586 (closest census date) so slope < 1 means that 
# population decreased over time

intercept <- as.numeric(coef(mod)[1])
slope <- as.numeric(coef(mod)[2]) #slope, close to 1 so we feel comfortable interpolating
```

Due to the high $r$ value (0.95) and the slope being close to 1, we feel comfortable linearly interpolating between the 1696 and 1586 population data to estimate each parish's population on January 1, 1629.

```{r interpolation}
# Per-parish interpolation
interpolated_1629 <- as.numeric(population_df$Population_1586) + (43 / 110) * (population_df$Population_1696 - population_df$Population_1586)

# Population-wide interpolation using linear regression results
# 1629 - 1586 = 43, 1696 - 1586 = 110, so to interpolate we multiply by 43/110
#interpolated_1629 <- (intercept - (1-slope) * population_pre$Population_1586) * 
#  43/110 + population_pre$Population_1586
```

Create population_specified, a dataframe containing only the populations of the parishes of interest.

```{r define_population_specified}
population <- data.frame(population_post$Parish, interpolated_1629 / 1000)
colnames(population) <- c("Parish","Population")
population_specified <- population[population$Parish %in% colnames(filled[,-1]),]
population_specified <- population_specified[match(colnames(filled[,-1]), population_specified$Parish),]
```

Now, we combine sestiere_pre and sestiere_post, the two dataframes that list the parishes within each sestiere (a sestiere is a larger geographical area that contains multiple parishes).

```{r combine_sestiere_dfs}
# Fix the parish names on both sestiere dataframes
sestiere_pre$Parish <- gsub("[[:space:]]", "\\.", sestiere_pre$Parish)
sestiere_pre$Parish <- sub("^S\\.", "S", sestiere_pre$Parish)
sestiere_post$Parish <- gsub("[[:space:]]", "\\.", sestiere_post$Parish)
sestiere_post$Parish <- sub("^S\\.", "S", sestiere_post$Parish)

# Arrange both sestiere dataframes by sestiere
sestiere_pre <- sestiere_pre %>% arrange(Sestiere, Parish)
sestiere_post <- sestiere_post %>% arrange(Sestiere, Parish)

# S. Marcilian and S.Simone.Piccolo is spelled S.Simeone.Piccolo in population_df, but Simeone Piccolo is not in filled because there was no data
#setdiff(sestiere_pre$Parish, population_df$Parish)

# Since we chose to interpolate using population_1586, we use sestiere_pre,
# which is taken from the same 1586 census, to ensure that the parish names match
sestiere_df <- sestiere_pre

# Match parishes in sestiere_df with those in filled
remove_vec <- c()
for(i in 1:length(sestiere_df$Parish)){
  if(!(sestiere_df$Parish[i] %in% colnames(filled[,-1]))){
    remove_vec <- c(remove_vec,i)
  }
}

if(length(remove_vec) > 0){
  sestiere_df <- sestiere_df[-remove_vec,]
}

# Define dataframes for each individual year
filled_1629 <- filled[year(filled$Date) == 1629,]
filled_1630 <- filled[year(filled$Date) == 1630,]
filled_1631 <- filled[year(filled$Date) == 1631,]
```

Finally, we figure out which parishes we will keep. The code below removes San Marco from our analysis because it served as a placeholder for deaths from drowning and those in prison, which resulted in an early spike in cases not representative of parish transmission. In our main analysis, we exclude it, but in the supplementary information, we include San Marco as a form of sensitivity analysis. The other two parishes we exclude are those which had no data during the primary outbreak window (9/1/30-12/31/30). 

```{r outlier_handling}
outliers <- c("S.Polo", "S.Stae", "S.Marco") #No cases between 9/1/30 and 12/31/30, outbreak time
filled_modified <- filled[, !(colnames(filled) %in% outliers)]
sestiere_df_modified <- sestiere_df[!(sestiere_df$Parish %in% outliers),]
population_specified_modified <- population_specified[!(population_specified$Parish %in% outliers),]
```

# Define Statistic-Calculating Functions

In this section, we define the functions that calculate the statistics of interest (entropy and KL divergence of the quantile function and PDF).

## Define PDF Approximation Function

For our histogram approximation of the PDF, we use Sturges's Rule to choose $K$, the number of bins, for $N$ observations ($K = \log_2(N)+1$).

```{r define_PDF_approx}
#Define a function that provides a histogram approximation of a PDF with K bins. Here, we use Sturges's Rule for K.
PDF_approx <- function(x, K = ceiling(log2(length(x))) + 1) {
  p <- table(cut(x, breaks = K)) / length(x)
  sigma2 <- sum((1:K) ^ 2 * p) - sum((1:K) * p) ^ 2
  p <- p[(p > 0)]
  return(p)
}
```

## Define Kullback-Leibler Divergence

To calculate Kullback-Leibler divergence between two discrete probability distributions P(x) and Q(x), we use the definition $\sum_{x \in \mathcal{X}} P(x) \log \frac{P(x)}{Q(x)}$. While this formulation implies that KL divergence is undefined when $P(x) = Q(x) = 0$, we [consider the contribution of the KL divergence at that point to be zero](https://en.wikipedia.org/wiki/Kullback%E2%80%93Leibler_divergence#Definition).

```{r define_KL}
#Next, we define Kullback-Leibler divergence between two discrete probability distributions (x and y), formatted as vectors. We add a small value epsilon to avoid dividing by zero.
KL_new <- function(x, y, epsilon = 1e-05){
  distance = 0
  pq_ratio = 0
  if(length(x) != length(y)){
    print("not same length")
    return(NA)
  }
  for(i in 1:length(x)){
    if(is.nan(x[i]) | is.nan(y[i]) | x[i] == 0){
      next
    }
    if(y[i] == 0){
      pq_ratio = x[i] / epsilon
    } else {
      pq_ratio = x[i] / y[i]
    }
    distance = distance + x[i] * log(pq_ratio)
  }
  return(distance)
}
  
```

For a discrete probability distribution $P(x)$ taking values in the set $\mathcal{X}$, we define the Shannon entropy $H$ to be $H = -\sum_{x \in \mathcal{X}} P(x) \ln P(x)$.

## Define Main Analaysis Function

Now, we define our main analysis function which returns the total number of deaths, Shannon entropy of the PDF, Shannon Entropy of the quantile function, negentropy of the PDF (KL divergence with a Gaussian), KL divergence of the PDF from a uniform, and KL divergence of the quantile from a discrete uniform at a given timestep. Later, we will run this function on the 999 resampled datasets generated during the bootstrapping process.

```{r define_entropy_procedure}
#---- 
# Documentation: entropy_procedure
# Usage: entropy_procedure(original_data, subset, indices, timestep, cumulative)
# @description: Return a list containing the sum of all cases (sum), 
# Shannon entropy of the PDF (shannonPDF), 
# Shannon entropy of the quantile function (shannonQuantile), 
# KL divergence of the PDF from a Gaussian distribution (negentPDF), 
# KL divergence of the PDF from a uniform distribution (uniformKL_PDF),
# and KL divergence of the quantile function from a uniform distribution (uniformKL_Quantile)
# calculated on mortality data in a specified subset of parishes on or up to a particular day

# Arguments/Options:
# @param original_data: a dataframe containing one variable for every parish of interest. Note that original_data does NOT get modified by subset, so original_data must already be subsetted if one only wishes to include a smaller selection of parishes.
# @param subset: a vector containing the indices of the parishes of interest in colnames(filled_modified).
# @param indices: a vector containing the indices generated by sampling with replacement. To analyze the original data, set indices = 1:ncol(original_data).
# @param timestep: an integer between 1 and nrow(filled_modified) corresponding to the day for which the aforementioned statistics should be calculated.
# @param cumulative: a boolean specifying whether deaths occurring on days prior to timestep should be included. If cumulative is set to FALSE, only deaths on day "timestep" will be considered. If cumulative is set to TRUE, deaths up to and including day "timestep" will be added up (DEFAULT: TRUE)

# @return: the list as described above
#----

entropy_procedure <- function(original_data,
           subset,
           indices,
           timestep,
           cumulative = TRUE) {
    boot_data <- original_data[, indices]
    # The proper index for pop_spec is subset - 1 because filled has a "Date"
    # column while population_specified_modified does not
    pop_spec <- population_specified_modified[subset - 1, ][indices, ]
    if (cumulative) {
      non_pop_corrected <- as.numeric(colSums(boot_data[1:timestep, ]))
    } else {
      non_pop_corrected <- as.numeric(boot_data[timestep, ])
    }
    
    # Case vector (cv): cases in each parish divided by parish population
    cv <- non_pop_corrected / (pop_spec$Population * 1000)
    cv <- sort(cv)
    
    # Normalized case vector (cv_norm): quantile function
    cv_norm <- cv / sum(cv)
    
    # Kernel (kern): histogram of mortality with # bins specified by Sturges's Rule
    kern <- as.numeric(PDF_approx(cv))
    
    # Normalized kernel (kern_norm): PMF
    kern_norm <- kern / sum(kern)
    
    shannon_PDF <- -1 * sum(kern[kern != 0] * log(kern[kern != 0]))
    shannon_quantile <-
      -1 * sum(cv_norm[cv_norm != 0] * log(cv_norm[cv_norm != 0]))
    
    # Use the msos negent function to calculate the negentropy of the PDF
    negent_PDF <- negent(cv) 

    uniform_KL_PDF <-
      as.numeric(KL_new(kern, rep(1 / length(kern), length(kern))))
    uniform_KL_quantile <-
      KL_new(cv_norm, rep(1 / length(cv_norm), length(cv_norm)))
    return(
      list(
        sum = sum(non_pop_corrected) / (sum(pop_spec$Population) * 1000),
        shannonPDF = shannon_PDF,
        shannonQuantile = shannon_quantile,
        negentPDF = negent_PDF,
        uniformKL_PDF = uniform_KL_PDF,
        uniformKL_Quantile = uniform_KL_quantile
      )
    )
  }
```

# Bootstrapping

In this section of code, we run bootstrapping (*n* = 999) to identify 95% confidence intervals for all of the statistics we calculate.

## Define Bootstrapping Functions

Now, we implement a generalized function to bootstrap over our entropy_procedure function. We will utilize this function in the subsequent functions bootstrapping over a particular dataset.

```{r define manual_boot_modified}
#----
# Documentation: manual_boot_modified
# Usage: manual_boot_modified(data, subset, N, max_ts, cumulative)
# @description: Returns a list containing six matrices formatted with dates as columns and resamples as rows. In order, the matrices contain: 
# the sum of all cases (sum_mat), 
# Shannon entropy of the PDF (shannonPDF_mat), 
# Shannon entropy of the quantile function (shannonQuantile_mat), 
# KL divergence of the PDF from a Gaussian distribution (negentPDF_mat), 
# KL divergence of the PDF from a uniform distribution (uniformKL_PDF_mat),
# and KL divergence of the quantile function from a uniform distribution (uniformKL_Quantile_mat)
# calculated on mortality data in a specified subset of parishes on or up to a particular day.

# Arguments/Options:
# @param data: a dataframe containing one variable for every parish of interest. Note that data does NOT get modified by the parameter "subset", so data must already be subsetted if one only wishes to include a smaller selection of parishes.
# @param subset: a vector containing the indices of the parishes of interest in colnames(filled_modified).
# @param N: an integer denoting the number of resamples to perform.
# @param max_ts: an integer corresponding to the final day for which the aforementioned statistics should be calculated.
# @param cumulative: a boolean specifying whether deaths occurring on days prior to timestep should be included. If cumulative is set to FALSE, the aforementioned calculations for a given day will only consider the deaths on that given day. If cumulative is set to TRUE, deaths up to and including that day will be summed.

# @return: the list as described above
#----

manual_boot_modified <-
  function(data, subset, N, max_ts, cumulative) {
    sum_mat <- matrix(NA, nrow = N, ncol = max_ts)
    shannonPDF_mat <- matrix(NA, nrow = N, ncol = max_ts)
    shannonQuantile_mat <- matrix(NA, nrow = N, ncol = max_ts)
    negentPDF_mat <- matrix(NA, nrow = N, ncol = max_ts)
    uniformKL_PDF_mat <- matrix(NA, nrow = N, ncol = max_ts)
    uniformKL_Quantile_mat <- matrix(NA, nrow = N, ncol = max_ts)
    # Iterate across N resamplings
    for (i in 1:N) {
      # Obtain a random sample of the data with replacement with length equal to the original data 
      sample_index <- sample(ncol(data), replace = TRUE)
      # Iterate across every day in the 3 year period
      for (t in 1:max_ts) {
        current_list <-
          entropy_procedure(original_data, subset, sample_index,
                                      t, cumulative)
        sum_mat[i, t] <- current_list$sum
        shannonPDF_mat[i, t] <- current_list$shannonPDF
        shannonQuantile_mat[i, t] <- current_list$shannonQuantile
        negentPDF_mat[i, t] <- current_list$negentPDF
        uniformKL_PDF_mat[i, t] <- current_list$uniformKL_PDF
        uniformKL_Quantile_mat[i, t] <-
          current_list$uniformKL_Quantile
      }
    }
    return(
      list(
        sum = sum_mat,
        shannonPDF = shannonPDF_mat,
        shannonQuantile = shannonQuantile_mat,
        negentPDF = negentPDF_mat,
        uniformKL_PDF = uniformKL_PDF_mat,
        uniformKL_Quantile = uniformKL_Quantile_mat
      )
    )
  }
```

## Identify 95% CI

Instead of finding the 97.5th and 2.5th percentile full runs, we find the 97.5th and 2.5th percentile values at each time point.

```{r define find_closest_run}
#----
# Documentation: find_closest_run
# Usage: find_closest_run(all_runs, target_percentile)
# @description: Returns in vector form the row of a specified dataframe with the nth lowest sum (n = target_percentile)

# Arguments/Options: 
# @param all_runs: a dataframe containing numeric values with one column for every time point and one row for every resampling
# @param target_percentile: an integer describing the index of the desired sorted vector all_runs. For example, if all_runs contains 1000 rows, the 97.5th percentile row selected by setting target_percentile = 975)

# @return: the numeric vector as described above
#----

find_closest_run <- function(all_runs, target_percentile) {
  all_runs_sum <- rowSums(all_runs)
  sorted_all_runs_sum <- sort(all_runs_sum)
  index <- match(sorted_all_runs_sum[target_percentile], all_runs_sum)
  return(as.numeric(all_runs[index,]))
}
```

## Per-sestiere analysis

Now, we define the analysis and bootstrapping functions for the per-sestiere analysis.

```{r define_sestiere_bootstrapping}
#----
# Documentation: shannonBySestiere
# Usage: shannonBySestiere(cumulative)
# @description: Returns a list of 6 lists (one for each sestiere). Each list contains six numerical vectors (one for each statistic) with entries for every day in the epidemic. 

# Arguments/Options:
# @param cumulative: a boolean specifying whether deaths occurring on days prior to timestep should be included. If cumulative is set to FALSE, the aforementioned calculations for a given day will only consider the deaths on that given day. If cumulative is set to TRUE, deaths up to and including that day will be summed (DEFAULT: TRUE)

# @return: the list as described above
#----
shannonBySestiere <- function(cumulative = TRUE) {
  # Initialize vectors
  sum_vec <- rep(0, length(filled_modified$Date))
  shannonPDF_vec <- rep(0, length(filled_modified$Date))
  shannonQuantile_vec <- rep(0, length(filled_modified$Date))
  negentPDF_vec <- rep(0, length(filled_modified$Date))
  uniformKL_PDF_vec <- rep(0, length(filled_modified$Date))
  uniformKL_Quantile_vec <- rep(0, length(filled_modified$Date))
  
  # Initialize the output list
  all_runs_list <-
    vector(mode = "list", length = length(unique(sestiere_df_modified$Sestiere)))
  
  for (s in unique(sestiere_df_modified$Sestiere)) {
    # Generate a vector containing the indices of the current sestiere's parishes
    sestiere_parishes <-
      match(sestiere_df_modified$Parish[sestiere_df_modified$Sestiere == s], colnames(filled_modified))
    for (i in 1:length(filled_modified$Date)) {
      current_list <-
        entropy_procedure(
          filled_modified[, sestiere_parishes], # Subset the data using sestiere_parishes
          subset = sestiere_parishes,
          seq(
            1,
            length(sestiere_parishes),
            length.out = length(sestiere_parishes)
          ),
          timestep = i,
          cumulative = cumulative
        )
      sum_vec[i] <- current_list$sum
      shannonPDF_vec[i] <- current_list$shannonPDF
      shannonQuantile_vec[i] <- current_list$shannonQuantile
      negentPDF_vec[i] <- current_list$negentPDF
      uniformKL_PDF_vec[i] <- current_list$uniformKL_PDF
      uniformKL_Quantile_vec[i] <- current_list$uniformKL_Quantile
    }
    # For each sestiere, store a list of the six statistic vectors in all_runs_list
    all_runs_list[[match(s, unique(sestiere_df_modified$Sestiere))]] <-
      list(
        sum = sum_vec,
        shannonPDF = shannonPDF_vec,
        shannonQuantile = shannonQuantile_vec,
        negentPDF = negentPDF_vec,
        uniformKL_PDF = uniformKL_PDF_vec,
        uniformKL_Quantile = uniformKL_Quantile_vec
      )
  }
  return(all_runs_list)
}

#----
# Documentation: shannonBySestiere_Bootstrapped
# Usage: shannonBySestiere_Bootstrapped(cumulative)
# @description: Returns a list containing 6 lists (one for each sestiere), each containing 5 matrices formatted with each statistic as a column and each day as a row. In order, the matrices contain: 
# the 5th percentile value of each statistic calculated separately on each day (b1), 
# the 95th percentile value of each statistic calculated separately on each day (b2), 
# the value of the 5th percentile resample for each statistic (closest_runs_025), 
# the value of the 5th percentile resample for each statistic (closest_runs_975), 
# calculated on mortality data in each sestiere on or up to a particular day.

# Arguments/Options:
# @param cumulative: a boolean specifying whether deaths occurring on days prior to timestep should be included. If cumulative is set to FALSE, the aforementioned calculations for a given day will only consider the deaths on that given day. If cumulative is set to TRUE, deaths up to and including that day will be summed (DEFAULT: TRUE)

# @return: the list as described above
#----

shannonBySestiere_Bootstrapped <- function(cumulative = TRUE) {
  num_dates <- length(filled_modified$Date)
  sestiere_list <- unique(sestiere_df_modified$Sestiere)
  num_sestieri <- length(sestiere_list)
  
  # Keep track of run time
  st <- Sys.time()
  
  # Initialize all_runs_list, which we will eventually return
  all_runs_list <- vector(mode = "list", length = length(sestiere_list))
  
  for (s in sestiere_list) {
    cat("Processing:", s, "\n")
    s_index <- match(s, sestiere_list)
    
    # Find the indices of the parishes in sestiere "s" in filled_modified
    sestiere_parishes <- match(sestiere_df_modified$Parish[sestiere_df_modified$Sestiere == s], colnames(filled_modified))
    
    # Subset the mortality data to only consider the parishes in sestiere "s"
    data_subset <- filled_modified[, sestiere_parishes]
    
    boot_mat <- manual_boot_modified(data = data_subset, 
                                       subset = sestiere_parishes, N = 999, 
                                       max_ts = num_dates, 
                                       cumulative = cumulative)
    
    #Initialize the upper and lower bound matrices
    b1_mat <- matrix(NA, ncol = length(boot_mat), nrow = num_dates)
    b2_mat <- matrix(NA, ncol = length(boot_mat), nrow = num_dates)
    
    closest_runs_025_mat <- matrix(NA, ncol = length(boot_mat), nrow = num_dates)
    closest_runs_975_mat <- matrix(NA,ncol = length(boot_mat), nrow = num_dates)
    #b1[, s_index] <- apply(boot_mat, 2, quantile, probs = 0.025, na.rm = TRUE)
    #b2[, s_index] <- apply(boot_mat, 2, quantile, probs = 0.975, na.rm = TRUE)
    for(j in 1:length(boot_mat)){
      closest_runs_025_mat[,j] <- find_closest_run(boot_mat[[j]], 25)
      closest_runs_975_mat[,j] <- find_closest_run(boot_mat[[j]], 975)
      for(i in 1:ncol(boot_mat[[j]])){
        sorted <- sort(as.numeric(boot_mat[[j]][,i]))
        b1_mat[i, j] <- sorted[25]
        b2_mat[i, j] <- sorted[975]
      }
    }
    all_runs_list[[match(s,sestiere_list)]] <- list(b1 = b1_mat, b2 = b2_mat, closest_runs_025 = closest_runs_025_mat, closest_runs_975 = closest_runs_975_mat)
  }
  ed <- Sys.time()
  print(ed - st)
  return(all_runs_list)
}
```

## Whole Dataset Analysis

Finally, we define the functions for the bootstrapping of the whole dataset.

```{r define_whole_dataset_analysis}
#----
# Documentation: whole_dataset_analysis
# Usage: whole_dataset_analysis(cumulative)
# @description: Returns a list of 6 vectors, each containing the value of a particular statistic for every day in the epidemic.

# Arguments/Options:
# @param cumulative: a boolean specifying whether deaths occurring on days prior to timestep should be included. If cumulative is set to FALSE, the aforementioned calculations for a given day will only consider the deaths on that given day. If cumulative is set to TRUE, deaths up to and including that day will be summed (DEFAULT: TRUE)

# @return: the list as described above
#----
whole_dataset_analysis <- function(cumulative = TRUE){
  # Initialize the vectors for each statistic
  sum_vec <- rep(0, length(filled_modified$Date)) # cumulative deaths
  shannonPDF_vec <- rep(0, length(filled_modified$Date)) # Shannon entropy of the PDF
  shannonQuantile_vec <- rep(0, length(filled_modified$Date)) # Shannon entropy of the quantile function
  negentPDF_vec <- rep(0, length(filled_modified$Date)) # KL divergence of the PDF from a Gaussian
  uniformKL_PDF_vec <- rep(0, length(filled_modified$Date)) # KL divergence of the PDF from a uniform
  uniformKL_Quantile_vec <- rep(0, length(filled_modified$Date)) # KL divergence of the quantile function from a uniform
  
  st <- Sys.time() # Record the starting time
  pb <- progress_bar$new(total = length(filled_modified$Date)) # Create a progress bar
  
  for(i in 1:length(filled_modified$Date)){
    current_list <- entropy_procedure(filled_modified[,-1], subset = 2:ncol(filled_modified), indices = 1:ncol(filled_modified[,-1]), timestep = i, cumulative)
    sum_vec[i] <- current_list$sum
    shannonPDF_vec[i] <- current_list$shannonPDF
    shannonQuantile_vec[i] <- current_list$shannonQuantile
    negentPDF_vec[i] <- current_list$negentPDF
    uniformKL_PDF_vec[i] <- current_list$uniformKL_PDF
    uniformKL_Quantile_vec[i] <- current_list$uniformKL_Quantile
    pb$tick()
  }
  ed <- Sys.time() # Record the final time
  print(ed - st) # Print the runtime
  return(list(sum = sum_vec, shannonPDF = shannonPDF_vec, shannonQuantile = shannonQuantile_vec, negentPDF = negentPDF_vec, uniformKL_PDF = uniformKL_PDF_vec, uniformKL_Quantile = uniformKL_Quantile_vec))
}

#----
# Documentation: whole_dataset_bootstrap
# Usage: whole_dataset_bootstrap(cumulative)
# @description: Returns a list of 4 lists. Lists b1 and b2 correspond to the 2.5th and 97.5th percentile values at every time step, respectively, while lists closest_run_025 and closest_runs_975 correspond to the values of the 2.5th and 97.5th percentile trials, respectively. 

# Arguments/Options:
# @param cumulative: a boolean specifying whether deaths occurring on days prior to timestep should be included. If cumulative is set to FALSE, the aforementioned calculations for a given day will only consider the deaths on that given day. If cumulative is set to TRUE, deaths up to and including that day will be summed (DEFAULT: TRUE)

# @return: the list as described above
#----
whole_dataset_bootstrap <- function(cumulative = TRUE){
  num_dates <- length(filled_modified$Date)
  b1 <- rep(0, num_dates) # 2.5%
  b2 <- rep(0, num_dates) # 97.5%
  boot_mat <- manual_boot_modified(data = filled_modified[,-1], subset = 2:ncol(filled_modified), N = 999, max_ts = num_dates, cumulative = cumulative)
  b1_list <- vector(mode = "list",length = length(boot_mat))
  b2_list <- vector(mode = "list",length = length(boot_mat))
  closest_runs_025_list <- vector(mode = "list", length(boot_mat))
  closest_runs_975_list <- vector(mode = "list", length(boot_mat))
    #b1[, s_index] <- apply(boot_mat, 2, quantile, probs = 0.025, na.rm = TRUE)
    #b2[, s_index] <- apply(boot_mat, 2, quantile, probs = 0.975, na.rm = TRUE)
  for(j in 1:length(boot_mat)){
    closest_025 <- find_closest_run(boot_mat[[j]], 25)
    closest_975 <- find_closest_run(boot_mat[[j]], 975)
    for(i in 1:ncol(boot_mat[[j]])){
      sorted <- sort(as.numeric(boot_mat[[j]][,i]))
      b1[i] <- sorted[25]
      b2[i] <- sorted[975]
    }
    b1_list[[j]] <- b1
    b2_list[[j]] <- b2
    closest_runs_025_list[[j]] <- closest_025
    closest_runs_975_list[[j]] <- closest_975
  }
  return(list(b1 = b1_list, b2 = b2_list, closest_runs_025 = closest_runs_025_list, closest_runs_975 = closest_runs_975_list))
}
```

## Run Bootstrap

Finally, we run our bootstrap!

```{r run_bootstrap, eval = FALSE}

set.seed(123) #unfortunately, we didn't set the seed ahead of time

stats_by_sestiere_boot_cumulative_50_2 <- shannonBySestiere_Bootstrapped(cumulative = TRUE) #w/ fixed negent
stats_by_sestiere_boot_instantaneous_50_2 <- shannonBySestiere_Bootstrapped(cumulative = FALSE)
stats_by_sestiere_cumulative_50_2 <- shannonBySestiere(cumulative = TRUE)
stats_by_sestiere_instantaneous_50_2 <- shannonBySestiere(cumulative = FALSE)
stats_of_whole_cumulative_50_2 <- whole_dataset_analysis(cumulative = TRUE) #default
stats_of_whole_instantaneous_50_2 <- whole_dataset_analysis(cumulative = FALSE)
stats_of_whole_boot_cumulative_50_2 <- whole_dataset_bootstrap(cumulative = TRUE)
stats_of_whole_boot_instantaneous_50_2 <- whole_dataset_bootstrap(cumulative = FALSE)

# Save our bootstrap values to a separate document
# saveRDS(stats_by_sestiere_boot_cumulative_50_2,"stats_by_sestiere_boot_cumulative_50_2.rds")
# saveRDS(stats_by_sestiere_boot_instantaneous_50_2,"stats_by_sestiere_boot_instantaneous_50_2.rds")
# saveRDS(stats_of_whole_boot_cumulative_50_2,"stats_of_whole_boot_cumulative_50_2.rds")
# saveRDS(stats_of_whole_boot_instantaneous_50_2,"stats_of_whole_boot_instantaneous_50_2.rds")
```

# Results

Now, we record our results. In this section, we focus on printing specific values of interest.

## Print Values on First and Last Day with 95% CIs

First, we print the true value, 2.5th percentile value, and 97.5th percentile value on the first and last days of the epidemic for each of our statistics calculated on the cumulative mortality data.

```{r print cumulative bootstrap results - first and last day values, eval = FALSE}
# Entropy of the quantile function on the first day
stats_of_whole_cumulative_50_2$shannonQuantile[1]
stats_of_whole_boot_cumulative_50_2$b1[[3]][1]
stats_of_whole_boot_cumulative_50_2$b2[[3]][1]

# Entropy of the quantile function on the final day
stats_of_whole_cumulative_50_2$shannonQuantile[nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b1[[3]][nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b2[[3]][nrow(filled_modified)]


# KL divergence of the quantile function on the first day from a uniform distribution
stats_of_whole_cumulative_50_2$uniformKL_Quantile[1]
stats_of_whole_boot_cumulative_50_2$b1[[6]][1]
stats_of_whole_boot_cumulative_50_2$b2[[6]][1]

# KL divergence of the quantile function on the final day from a uniform distribution
stats_of_whole_cumulative_50_2$uniformKL_Quantile[nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b1[[6]][nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b2[[6]][nrow(filled_modified)]


# Entropy of the PDF on the first day
stats_of_whole_cumulative_50_2$shannonPDF[1]
stats_of_whole_boot_cumulative_50_2$b1[[2]][1]
stats_of_whole_boot_cumulative_50_2$b2[[2]][1]

# Entropy of the PDF on the final day
stats_of_whole_cumulative_50_2$shannonPDF[nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b1[[2]][nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b2[[2]][nrow(filled_modified)]


# KL divergence of the PDF on the first day from a Gaussian distribution
stats_of_whole_cumulative_50_2$negentPDF[1]
stats_of_whole_boot_cumulative_50_2$b1[[4]][1]
stats_of_whole_boot_cumulative_50_2$b2[[4]][1]

# KL divergence of the PDF on the final day from a Gaussian distribution
stats_of_whole_cumulative_50_2$negentPDF[nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b1[[4]][nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b2[[4]][nrow(filled_modified)]


# KL divergence of the PDF on the first day from a uniform distribution 
stats_of_whole_cumulative_50_2$uniformKL_PDF[1]
stats_of_whole_boot_cumulative_50_2$b1[[5]][1]
stats_of_whole_boot_cumulative_50_2$b2[[5]][1]

# KL divergence of the PDF on the final day from a uniform distribution
stats_of_whole_cumulative_50_2$uniformKL_PDF[nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b1[[5]][nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$b2[[5]][nrow(filled_modified)]

# Same as previous, but using the "closest runs" methodology to find the 2.5th and 97.5th percentile values
stats_of_whole_boot_cumulative_50_2$closest_runs_025[[5]][nrow(filled_modified)]
stats_of_whole_boot_cumulative_50_2$closest_runs_975[[5]][nrow(filled_modified)]
```

Second, we print the extrema of each statistic (minima for KL divergence, maxima for entropy) calculated on the cumulative mortality data, the day on which they occurred, and the 2.5th and 97.5th percentile values on that day.

```{r print cumulative bootstrap results - extrema, eval = FALSE}
# Minimum KL divergence of the PDF from a uniform distribution
min(stats_of_whole_cumulative_50_2$uniformKL_PDF)
min_index <- match(min(stats_of_whole_cumulative_50_2$uniformKL_PDF),stats_of_whole_cumulative_50_2$uniformKL_PDF)
filled$Date[min_index]
stats_of_whole_cumulative_50_2$uniformKL_PDF[min_index]
stats_of_whole_boot_cumulative_50_2$b1[[5]][min_index]
stats_of_whole_boot_cumulative_50_2$b2[[5]][min_index]

# Minimum KL divergence of the PDF from a Gaussian distribution
min(stats_of_whole_cumulative_50_2$negentPDF)
min_index <- match(min(stats_of_whole_cumulative_50_2$negentPDF),stats_of_whole_cumulative_50_2$negentPDF)
filled$Date[min_index]
stats_of_whole_cumulative_50_2$negentPDF[min_index]
stats_of_whole_boot_cumulative_50_2$b1[[4]][min_index]
stats_of_whole_boot_cumulative_50_2$b2[[4]][min_index]

# Minimum KL divergence of the quantile function from a uniform distribution
min(stats_of_whole_cumulative_50_2$uniformKL_Quantile)
min_index <- match(min(stats_of_whole_cumulative_50_2$uniformKL_Quantile),stats_of_whole_cumulative_50_2$uniformKL_Quantile)
filled$Date[min_index]
stats_of_whole_cumulative_50_2$uniformKL_Quantile[min_index]
stats_of_whole_boot_cumulative_50_2$b1[[6]][min_index]
stats_of_whole_boot_cumulative_50_2$b2[[6]][min_index]

# Maximum entropy of the quantile function
max(stats_of_whole_cumulative_50_2$shannonQuantile)
max_index <- match(max(stats_of_whole_cumulative_50_2$shannonQuantile),stats_of_whole_cumulative_50_2$shannonQuantile)
filled$Date[max_index]
stats_of_whole_cumulative_50_2$shannonQuantile[max_index]
stats_of_whole_boot_cumulative_50_2$b1[[3]][max_index]
stats_of_whole_boot_cumulative_50_2$b2[[3]][max_index]

# Maximum entropy of the PDF
max(stats_of_whole_cumulative_50_2$shannonPDF)
max_index <- match(max(stats_of_whole_cumulative_50_2$shannonPDF),stats_of_whole_cumulative_50_2$shannonPDF)
filled$Date[max_index]
stats_of_whole_cumulative_50_2$shannonPDF[max_index]
stats_of_whole_boot_cumulative_50_2$b1[[2]][max_index]
stats_of_whole_boot_cumulative_50_2$b2[[2]][max_index]
```

Then, we do the same for our statistics calculated on the instantaneous mortality data.

First and last day values:

```{r print instantaneous_bootstrap results - first and last day values, eval = FALSE}
# Entropy of the quantile function on the first day
stats_of_whole_instantaneous_50_2$shannonQuantile[1]
stats_of_whole_boot_instantaneous_50_2$b1[[3]][1]
stats_of_whole_boot_instantaneous_50_2$b2[[3]][1]

# Entropy of the quantile function on the final day
stats_of_whole_instantaneous_50_2$shannonQuantile[nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b1[[3]][nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b2[[3]][nrow(filled_modified)]


# KL divergence of the quantile function on the first day from a uniform distribution
stats_of_whole_instantaneous_50_2$uniformKL_Quantile[1]
stats_of_whole_boot_instantaneous_50_2$b1[[6]][1]
stats_of_whole_boot_instantaneous_50_2$b2[[6]][1]

# KL divergence of the quantile function on the final day from a uniform distribution
stats_of_whole_instantaneous_50_2$uniformKL_Quantile[nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b1[[6]][nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b2[[6]][nrow(filled_modified)]


# Entropy of the PDF on the first day
stats_of_whole_instantaneous_50_2$shannonPDF[1]
stats_of_whole_boot_instantaneous_50_2$b1[[2]][1]
stats_of_whole_boot_instantaneous_50_2$b2[[2]][1]

# Entropy of the PDF on the final day
stats_of_whole_instantaneous_50_2$shannonPDF[nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b1[[2]][nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b2[[2]][nrow(filled_modified)]


# KL divergence of the PDF on the first day from a Gaussian distribution
stats_of_whole_instantaneous_50_2$negentPDF[1]
stats_of_whole_boot_instantaneous_50_2$b1[[4]][1]
stats_of_whole_boot_instantaneous_50_2$b2[[4]][1]

# KL divergence of the PDF on the final day from a Gaussian distribution
stats_of_whole_instantaneous_50_2$negentPDF[nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b1[[4]][nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b2[[4]][nrow(filled_modified)]


# KL divergence of the PDF on the first day from a uniform distribution
stats_of_whole_instantaneous_50_2$uniformKL_PDF[1]
stats_of_whole_boot_instantaneous_50_2$b1[[5]][1]
stats_of_whole_boot_instantaneous_50_2$b2[[5]][1]

# KL divergence of the PDF on the final day from a uniform distribution
stats_of_whole_instantaneous_50_2$uniformKL_PDF[nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b1[[5]][nrow(filled_modified)]
stats_of_whole_boot_instantaneous_50_2$b2[[5]][nrow(filled_modified)]
```

Extrema:

```{r print instantaneous bootstrap results - extrema, eval = FALSE}
# KL divergence of the PDF from a uniform distribution
min(stats_of_whole_instantaneous_50_2$uniformKL_PDF)
min_index <- match(min(stats_of_whole_instantaneous_50_2$uniformKL_PDF),stats_of_whole_instantaneous_50_2$uniformKL_PDF)
filled$Date[min_index]
stats_of_whole_instantaneous_50_2$uniformKL_PDF[min_index]
stats_of_whole_boot_instantaneous_50_2$b1[[5]][min_index]
stats_of_whole_boot_instantaneous_50_2$b2[[5]][min_index]

# KL divergence of the PDF from a Gaussian distribution
min(stats_of_whole_instantaneous_50_2$negentPDF)
min_index <- match(min(stats_of_whole_instantaneous_50_2$negentPDF),stats_of_whole_instantaneous_50_2$negentPDF)
filled$Date[min_index]
stats_of_whole_instantaneous_50_2$negentPDF[min_index]
stats_of_whole_boot_instantaneous_50_2$b1[[4]][min_index]
stats_of_whole_boot_instantaneous_50_2$b2[[4]][min_index]

# KL divergence of the quantile function from a uniform distribution
min(stats_of_whole_instantaneous_50_2$uniformKL_Quantile)
min_index <- match(min(stats_of_whole_instantaneous_50_2$uniformKL_Quantile),stats_of_whole_instantaneous_50_2$uniformKL_Quantile)
filled$Date[min_index]
stats_of_whole_instantaneous_50_2$uniformKL_Quantile[min_index]
stats_of_whole_boot_instantaneous_50_2$b1[[6]][min_index]
stats_of_whole_boot_instantaneous_50_2$b2[[6]][min_index]

# Maximum entropy of the quantile function
max(stats_of_whole_instantaneous_50_2$shannonQuantile)
max_index <- match(max(stats_of_whole_instantaneous_50_2$shannonQuantile),stats_of_whole_instantaneous_50_2$shannonQuantile)
filled$Date[max_index]
stats_of_whole_instantaneous_50_2$shannonQuantile[max_index]
stats_of_whole_boot_instantaneous_50_2$b1[[3]][max_index]
stats_of_whole_boot_instantaneous_50_2$b2[[3]][max_index]

# Maximum entropy of the PDF
max(stats_of_whole_instantaneous_50_2$shannonPDF)
max_index <- match(max(stats_of_whole_instantaneous_50_2$shannonPDF),stats_of_whole_instantaneous_50_2$shannonPDF)
filled$Date[max_index]
stats_of_whole_instantaneous_50_2$shannonPDF[max_index]
stats_of_whole_boot_instantaneous_50_2$b1[[2]][max_index]
stats_of_whole_boot_instantaneous_50_2$b2[[2]][max_index]
```

## Descriptive Statistics

Next, we print descriptive statistics for our dataset, such as the median and mean parish population.

```{r print_descriptive_statistics}
# Number of parishes (first column is dates)
ncol(filled_modified[,-1])

# Total population on January 1, 1629
sum(population_specified_modified$Population*1000)

# Median parish population on January 1, 1629
median(population_specified_modified$Population*1000)

# Mean parish population on January 1, 1629
mean(population_specified_modified$Population*1000)

# Standard deviation in parish population on January 1, 1629
sd(population_specified_modified$Population*1000)

# Total deaths
sum(rowSums(filled_modified[,-1]))

# Mean cumulative parish mortality rate on December 31, 1631
mean(colSums(filled_modified[,-1])/(population_specified_modified$Population*1000))

# Standard deviation in cumulative parish mortality rate on December 31, 1631
sd(colSums(filled_modified[,-1])/(population_specified_modified$Population*1000))
```

## Effective Number of Parishes

Next, we validate our results against the effective number of parties $\left(\frac{1}{\sum_{i}^{n}p_i^2}\right)$, which has a more direct interpretation.

```{r effective number, eval = FALSE}
#Effective number and log of non-zeroes
non_zeroes_cumulative <- rep(0,length(filled$Date))
eff_number_cumulative <- rep(0,length(filled$Date))
non_zeroes_instantaneous <- rep(0,length(filled$Date))
eff_number_instantaneous <- rep(0,length(filled$Date))

for(i in 1:length(filled$Date)){
  case_vec_cumulative <- as.numeric(colSums(filled_modified[1:i,-1]))/(population_specified_modified$Population*1000)
  case_vec_cumulative_norm <- case_vec_cumulative/sum(case_vec_cumulative)
  case_vec_instantaneous <- as.numeric(filled_modified[i,-1])/(population_specified_modified$Population*1000)
  case_vec_instantaneous_norm <- case_vec_instantaneous/sum(case_vec_instantaneous)
  case_vec_cumulative_norm_nonzeroes <- case_vec_cumulative_norm[case_vec_cumulative_norm != 0]
  non_zeroes_cumulative[i] <- length(case_vec_cumulative_norm_nonzeroes)
  eff_number_cumulative[i] <- 1/sum(case_vec_cumulative_norm_nonzeroes^2)
  case_vec_instantaneous_norm_nonzeroes <- case_vec_instantaneous_norm[case_vec_instantaneous_norm != 0]
  non_zeroes_instantaneous[i] <- length(case_vec_instantaneous_norm_nonzeroes)
  eff_number_instantaneous[i] <- 1/sum(case_vec_instantaneous_norm_nonzeroes^2)
}

# Inverse Simpson's Index is Diversity with q = 2 while the Shannon Index is Diversity with q = 1
# plot.ts(log(non_zeroes_cumulative),ylim=c(2,4),col="red")
# lines(stats_of_whole_cumulative_50_2$shannonQuantile,col="black")
# plot.ts(log(non_zeroes_instantaneous),ylim=c(0,4),col="red")
# lines(stats_of_whole_instantaneous_50_2$shannonQuantile,col="black")
# plot.ts(log(eff_number_cumulative),ylim=c(2,4),col="red")
# lines(stats_of_whole_cumulative_50_2$shannonQuantile,col="black")
# plot.ts(log(eff_number_instantaneous),ylim=c(0,4),col="red")
# lines(stats_of_whole_instantaneous_50_2$shannonQuantile,col="black")
```

# Figures

Finally, we plot our results!

## Figure 2

First, we create Figure 2, a grid plot consisting of:

1.  Shannon entropy of the quantile function of cumulative mortality for every day in the epidemic along with corresponding 95% confidence intervals
2.  KL divergence of the quantile function of cumulative mortality from a uniform distribution for every day in the epidemic along with corresponding 95% confidence

Note that if we included the parish of S. Marco above, this will instead generate Supplementary Figure S4.

```{r plot results - figure 2, eval = FALSE}
#1: sum, 2: shannonPDF, 3: shannonQuantile, 4: negentPDF, 5: uniformKL_PDF, 6: uniformKL_Quantile
# For x-axis, two options: time or incidence ("sum")

# ----
# Documentation: Figure 2
# Goal: create a grid plot of shannonQuantile and uniformKL_Quantile including 95% CIs
# ----

# Initialize dataframe with 2.5th percentile value, 97.5th percentile value, and original value of uniformKL_Quantile
plot_df <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_cumulative_50_2$b1[[6]],
      stats_of_whole_boot_cumulative_50_2$b2[[6]],
      stats_of_whole_cumulative_50_2$uniformKL_Quantile
    )
  )
colnames(plot_df) <- c("b1", "b2", "observed")

plot1 <- ggplot(data = plot_df, aes(x = filled$Date, y = observed)) +
  geom_line() +
  geom_ribbon(aes(ymin = b1, ymax = b2), alpha = 0.5) + #gray out 95% CI
  geom_line(aes(y = rep(0, length(filled$Date))), col = "red") + #create red line indicating minimum possible KL divergence
  coord_cartesian(ylim = c(0, 0.35)) +
  # coord_cartesian(ylim = c(0, 0.5))+ #S. Marco
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous() +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Negentropy (nats)", x = "Date", title = "Negentropy")

# Initialize dataframe with 2.5th percentile value, 97.5th percentile value, and original value of shannonQuantile
plot_df_2 <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_cumulative_50_2$b1[[3]],
      stats_of_whole_boot_cumulative_50_2$b2[[3]],
      stats_of_whole_cumulative_50_2$shannonQuantile,
      eff_number_cumulative
    )
  )
colnames(plot_df_2) <- c("b1", "b2", "observed", "eff")

plot2 <- ggplot(data = plot_df_2, aes(x = filled$Date, y = observed)) +
  geom_line() +
  #geom_line(aes(y=rep(log(ncol(filled_modified[,-1])),length(filled$Date))),linetype="dashed")+
  geom_line(aes(y = rep(log(
    ncol(filled_modified[, -1])
  ), length(filled$Date))), col = "red") +
  geom_ribbon(aes(ymin = b1, ymax = b2), alpha = 0.5) +
  #geom_line(aes(y=log(eff)),col="red")+
  # coord_cartesian(ylim=c(3.5,3.95))+ #S. Marco
  coord_cartesian(ylim = c(3.5, 3.9)) +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous(
    name = "Entropy (nats)",
    sec.axis = sec_axis( #create second y-axis for effective number of parishes
      trans =  ~ .,
      name = "Effective Number",
      labels = function(x)
        scales::comma(round(exp(x)), accuracy = 1)
    )
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Entropy (nats)", x = "Date", title = "Entropy")

# Order such that entropy graph is on the left and KL divergence graph is on the right
plot_ls <- list(plot2, plot1)

# Create grid
untitled_grid <-
  plot_grid(plotlist = plot_ls,
            labels = c("a", "b"),
            label_size = 12)
# now add the title
title <- ggdraw() +
  draw_label(
    "Quantile Function of Cumulative Mortality",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(# add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7))
plot_grid(title,
          untitled_grid,
          ncol = 1,
          # rel_heights values control vertical title margins
          rel_heights = c(0.1, 1))

# Save figure

#ggsave("CumulativeQuantileGrid_Final.png",width=2100,height=1200,units="px")
#ggsave("CumulativeQuantileGrid_Final.jpg",width=2100,height=1200,units="px")
#ggsave("~/VenetianPlague_Entropy/Figures/Figure2.png",width=2100,height=1200,units="px")
#ggsave("~/VenetianPlague_Entropy/Figures/Figure2.jpg",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure2.png",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure2.jpg",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/SMarcoCumulativeQuantile.png",width=2100,height=1200,units="px") #Supp Fig S4
# ggsave("~/VenetianPlague_Entropy/SMarcoCumulativeQuantile.jpg",width=2100,height=1200,units="px")
```

## Figure 3

Second, we create Figure 3, a grid plot consisting of:

1. Shannon entropy of the PDF function of cumulative mortality for every day in the epidemic along with corresponding 95% confidence intervals
2. KL divergence of the PDF of cumulative mortality from a Gaussian distribution with the same mean and variance as the data for every day in the epidemic along with corresponding 95% confidence intervals

Note that if we included the parish of S. Marco above, this will instead generate Supplementary Figure S5.

```{r plot results - figure 3, eval = FALSE}
#1: sum, 2: shannonPDF, 3: shannonQuantile, 4: negentPDF, 5: uniformKL_PDF, 6: uniformKL_Quantile

# ----
# Documentation: Figure 3
# Goal: create a grid plot of shannonPDF and negentPDF including 95% CIs
# ----

# Initialize dataframe containing 2.5th percentile values, 97.5th percentile values, and observed values of the entropy of the PDF
plot_df <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_cumulative_50_2$b1[[2]],
      stats_of_whole_boot_cumulative_50_2$b2[[2]],
      stats_of_whole_cumulative_50_2$shannonPDF
    )
  )
colnames(plot_df) <- c("b1", "b2", "observed")

plot1 <- ggplot(data = plot_df, aes(x = filled$Date, y = observed)) +
  geom_line() +
  #geom_line(aes(y=rep(log(ceiling(log2(ncol(filled_modified[,-1])))+1),length(filled$Date))),linetype="dashed")+
  geom_line(aes(y = rep(log(ceiling(
    log2(ncol(filled_modified[, -1])) #ceiling(log2(N) + 1) is the number of bins under Sturges' Rule, but this should be the entropy of the Gaussian with the same mean and variance as the data on the final day
  ) + 1), length(filled$Date))), col = "red") +
  geom_ribbon(aes(ymin = b1, ymax = b2), alpha = 0.5) +
  coord_cartesian(ylim = c(1, 2)) +
  # coord_cartesian(ylim=c(0,2))+ #S. Marco
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous() +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Entropy (nats)", x = "Date", title = "Entropy")

# Initialize dataframe containing 2.5th percentile values, 97.5th percentile values, and observed values of the negentropy of the PDF
plot_df_2 <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_cumulative_50_2$b1[[4]],
      stats_of_whole_boot_cumulative_50_2$b2[[4]],
      stats_of_whole_cumulative_50_2$negentPDF
    )
  )
colnames(plot_df_2) <- c("b1", "b2", "observed")

plot2 <- ggplot(data = plot_df_2, aes(x = filled$Date, y = observed)) +
  geom_line() +
  #geom_line(aes(y=funny_vec))+
  geom_ribbon(aes(ymin = b1, ymax = b2), alpha = 0.5) +
  geom_line(aes(y = rep(0, length(filled$Date))), col = "red") +
  coord_cartesian(ylim = c(0, 0.45)) +
  # coord_cartesian(ylim=c(0,1.5))+ #S. Marco
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous() +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  #labs(y=expression(paste(D[KL]," (nats)")),x="Date",title="Negentropy")
  labs(y = "Negentropy (nats)", x = "Date", title = "Negentropy")

# Experimental Analysis: KL divergence of PDF from uniform (excluded in final paper)
# plot_df_3 <- as.data.frame(cbind(stats_of_whole_boot_cumulative_50_2$b1[[5]],stats_of_whole_boot_cumulative_50_2$b2[[5]],stats_of_whole_cumulative_50_2$uniformKL_PDF))
# colnames(plot_df_3) <- c("b1","b2","observed")
# plot3 <- ggplot(data=plot_df_3,aes(x=filled$Date,y=observed))+
#   geom_line()+
#   geom_ribbon(aes(ymin=b1,ymax=b2),alpha=0.5)+
#   coord_cartesian(ylim=c(0,1))+
#   scale_x_date(date_breaks="12 months",date_labels="%m-%Y")+
#   scale_y_continuous()+
#   theme(plot.background=element_rect(fill="white"),axis.title=element_text(size=10))+
#   labs(y="KL Divergence",x="Date",title="KL Divergence from Uniform")
# plot_ls <- list(plot1,plot2,plot3)
# untitled_grid <- plot_grid(plotlist=plot_ls,labels=c("A","B","C"),label_size=12)

# Order such that entropy graph is on the left and KL divergence graph is on the right
plot_ls <- list(plot1, plot2)

# Create the grid 
untitled_grid <-
  plot_grid(plotlist = plot_ls,
            labels = c("a", "b"),
            label_size = 12)

# Add the title
title <- ggdraw() +
  draw_label(
    "PDF of Cumulative Mortality",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(# add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7))

plot_grid(title,
          untitled_grid,
          ncol = 1,
          # rel_heights values control vertical title margins
          rel_heights = c(0.1, 1))

# Save the plot

#ggsave("cumulative_50_2_2PDFGrid5.png",width=2100,height=150_20,units="px")
#ggsave("CumulativePDFGrid_Final.png",width=2100,height=1200,units="px")
#ggsave("CumulativePDFGrid_Final.jpg",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures/Figure3.png",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures/Figure3.jpg",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure3.png",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure3.jpg",width=2100,height=1200,units="px")
# ggsave("~/VenetianPlague_Entropy/SMarcoCumulativePDF.png",width=2100,height=1200,units="px") # Supp Fig S5
# ggsave("~/VenetianPlague_Entropy/SMarcoCumulativePDF.jpg",width=2100,height=1200,units="px")
```

## Supplementary Figure S2

Third, we create Supplementary Figure S2, a grid plot containing:

1.  Shannon entropy of the quantile function of instantaneous mortality for every day in the epidemic along with corresponding 95% confidence intervals
2.  KL divergence of the quantile function of instantaneous mortality from a uniform distribution for every day in the epidemic along with corresponding 95% confidence intervals

```{r plot results - supplementary figure s2, eval = FALSE}
#1: sum, 2: shannonPDF, 3: shannonQuantile, 4: negentPDF, 5: uniformKL_PDF, 6: uniformKL_Quantile
#can plot against incidence ("sum") if you want

#----
# Documentation: Supplementary Figure S2
# Goal: Grid plot with instantaneous shannonQuantile (left) and instantaneous uniformKL_PDF (right)
#----

# Initialize dataframe containing 2.5th percentile, 97.5th percentile, and observed values of KL divergence of the quantile function from a uniform distribution
plot_df <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_instantaneous_50_2$b1[[6]],
      stats_of_whole_boot_instantaneous_50_2$b2[[6]],
      stats_of_whole_instantaneous_50_2$uniformKL_Quantile
    )
  )
colnames(plot_df) <- c("b1", "b2", "observed")

plot1 <- ggplot(data = plot_df, aes(x = filled$Date, y = observed)) +
  geom_line(lwd = 0.25) +
  geom_ribbon(aes(ymin = b1, ymax = b2), alpha = 0.5) +
  geom_line(aes(y = rep(0, length(filled$Date))), col = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous() +
  expand_limits(y = c(0, 2)) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10)
  ) +
  #labs(y=expression(D[K*L]),x="Date",title="Negentropy")
  labs(y = "Negentropy (nats)", x = "Date", title = "Negentropy")

# Initialize dataframe containing 2.5th percentile, 97.5th percentile, and original values of entropy of the quantile function
plot_df_2 <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_instantaneous_50_2$b1[[3]],
      stats_of_whole_boot_instantaneous_50_2$b2[[3]],
      stats_of_whole_instantaneous_50_2$shannonQuantile,
      log(non_zeroes_instantaneous)
    )
  )
colnames(plot_df_2) <- c("b1", "b2", "observed", "nonzeroes")

plot2 <- ggplot(data = plot_df_2, aes(x = filled$Date, y = observed)) +
  #geom_line(aes(y=nonzeroes),lwd=0.25,col="red",alpha=0.75)+ #optional, plots number of parishes with non-zero deaths
  geom_line(lwd = 0.25) +
  geom_line(aes(y = rep(log(
    ncol(filled_modified[, -1])
  ), length(filled$Date))), col = "red") +
  geom_ribbon(aes(ymin = b1, ymax = b2), alpha = 0.5) +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous() +
  expand_limits(y = c(2, 4)) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10)
  ) +
  labs(y = "Entropy (nats)", x = "Date", title = "Entropy")

# Order such that entropy graph is on the left and KL divergence graph is on the right
plot_ls <- list(plot2, plot1)
untitled_grid <-
  plot_grid(plotlist = plot_ls,
            labels = c("a", "b"),
            label_size = 12)

# Add the title
title <- ggdraw() +
  draw_label(
    "Quantile Function of Instantaneous Mortality",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(# add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7))

# Create the grid 
plot_grid(title,
          untitled_grid,
          ncol = 1,
          # rel_heights values control vertical title margins
          rel_heights = c(0.1, 1))

# Save the figure

#ggsave("InstantaneousQuantileGrid_Final_2.png",width=2100,height=1200,units="px")
#ggsave("InstantaneousQuantileGrid_Final_2.jpg",width=2100,height=1200,units="px")
#ggsave("InstantaneousQuantileGrid_Final.jpg",width=2100,height=1200,units="px")
#ggsave("InstantaneousQuantileGrid_Final_3.jpg",width=2100,height=1200,units="px")
# ggsave("InstantaneousQuantileGrid_Final_4.jpg",width=2100,height=1200,units="px") #Supp Fig S2
```

## Supplementary Figure S3

Fourth, we create Supplementary Figure S3, a grid plot containing:

1.  Shannon entropy of the PDF function of instantaneous mortality for every day in the epidemic along with corresponding 95% confidence intervals
2.  KL divergence of the PDF of instantaneous mortality from a Gaussian distribution with the same mean and variance as the data for every day in the epidemic along with corresponding 95% confidence intervals

```{r plot results - supplementary figure s3, eval = FALSE}
#1: sum, 2: shannonPDF, 3: shannonQuantile, 4: negentPDF, 5: uniformKL_PDF, 6: uniformKL_Quantile

#----
# Documentation: Supplementary Figure S3
# Goal: Grid plot of shannonPDF (left) and negentPDF (right) for instantaneous mortality data
#----

# Initialize dataframe containing 2.5th percentile, 97.5th percentile, and original values of the entropy of the PDF
plot_df <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_instantaneous_50_2$b1[[2]],
      stats_of_whole_boot_instantaneous_50_2$b2[[2]],
      stats_of_whole_instantaneous_50_2$shannonPDF
    )
  )
colnames(plot_df) <- c("b1", "b2", "observed")

plot1 <- ggplot(data = plot_df, aes(x = filled$Date, y = observed)) +
  geom_ribbon(aes(
    x = filled$Date,
    ymin = b1,
    ymax = b2
  ), alpha = 0.5) +
  geom_line(lwd = 0.25) +
  geom_line(aes(y = rep(log(ceiling(
    log2(ncol(filled_modified[, -1]))
  ) + 1), length(filled$Date))), col = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous() +
  expand_limits(y = c(0, 2)) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Entropy (nats)", x = "Date", title = "Entropy")

# Initialize dataframe containing 2.5th percentile, 97.5th percentile, and original values of the negentropy of the PDF
plot_df_2 <-
  as.data.frame(
    cbind(
      stats_of_whole_boot_instantaneous_50_2$b1[[4]],
      stats_of_whole_boot_instantaneous_50_2$b2[[4]],
      stats_of_whole_instantaneous_50_2$negentPDF
    )
  )
colnames(plot_df_2) <- c("b1", "b2", "observed")

plot2 <- ggplot(data = plot_df_2, aes(x = filled$Date, y = observed)) +
  geom_ribbon(aes(
    x = filled$Date,
    ymin = b1,
    ymax = b2
  ), alpha = 0.5) +
  geom_line(lwd = 0.25) +
  geom_line(aes(y = rep(0, length(filled$Date))), col = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  scale_y_continuous() +
  coord_cartesian(ylim = c(0, 1.5)) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Negentropy (nats)", x = "Date", title = "Negentropy")

# Experimental Analysis: include KL divergence of the PDF from a uniform distribution (excluded in final paper)
# plot_df_3 <- as.data.frame(cbind(stats_of_whole_boot_instantaneous_50_2$b1[[5]],stats_of_whole_boot_instantaneous_50_2$b2[[5]],stats_of_whole_instantaneous_50_2$uniformKL_PDF))
# colnames(plot_df_3) <- c("b1","b2","observed")
# plot3 <- ggplot(data=plot_df_3,aes(x=filled$Date,y=observed))+
#   geom_ribbon(aes(x=filled$Date,ymin=b1,ymax=b2),alpha=0.5)+
#   geom_line(lwd=0.25)+
#   scale_x_date(date_breaks="12 months",date_labels="%m-%Y")+
#   scale_y_continuous()+
#   expand_limits(y=c(0,1.8))+
#   theme(plot.background=element_rect(fill="white"),axis.title=element_text(size=10))+
#   labs(y="KL Divergence",x="Date",title="KL Divergence from Uniform")

# Order plots such that entropy graph is on the left and KL divergence graph is on the right
plot_ls <- list(plot1, plot2)
untitled_grid <-
  plot_grid(plotlist = plot_ls,
            labels = c("a", "b"),
            label_size = 12)

# Add the title
title <- ggdraw() +
  draw_label(
    "PDF of Instantaneous Mortality",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(# add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7))

# Create the grid
plot_grid(title,
          untitled_grid,
          ncol = 1,
          # rel_heights values control vertical title margins
          rel_heights = c(0.1, 1))

# Save the figure

#ggsave("InstantaneousPDFGrid_Final_2.png",width=2100,height=1200,units="px")
#ggsave("InstantaneousPDFGrid_Final_2.jpg",width=2100,height=1200,units="px")
#ggsave("InstantaneousPDFGrid_Final.jpg",width=2100,height=1200,units="px")
#ggsave("InstantaneousPDFGrid_Final_3.jpg",width=2100,height=1200,units="px")
# ggsave("InstantaneousPDFGrid_Final_4.jpg",width=2100,height=1200,units="px") # Supp Fig S3
```

## Figure 4

Fifth, we create Figure 4, a plot of Shannon entropy of the quantile functions for each sestiere's cumulative mortality data along with the 95% confidence intervals generated from bootstrapping.

```{r plot results - figure 4, eval = FALSE}
#----
# Documentation: Figure 4
# Goal: Time series plot of shannonQuantile of each sestiere, differentiated via colors
# ----

stat_of_interest <-
  3 #1: sum, 2: shannonPDF, 3: shannonQuantile, 4: negentPDF, 5: uniformKL_PDF, 6: uniformKL_Quantile
b1_plot_mat <-
  matrix(NA, nrow = nrow(filled_modified), ncol = 6) #6 = number of stats, not sestieri
b2_plot_mat <- matrix(NA, nrow = nrow(filled_modified), ncol = 6)
observed_plot_mat <- matrix(NA, nrow = nrow(filled_modified), ncol = 6)

for (i in 1:6) { #6 = number of sestieri
  b1_plot_mat[, i] <-
    stats_by_sestiere_boot_cumulative_50_2[[i]]$b1[, stat_of_interest]
  b2_plot_mat[, i] <-
    stats_by_sestiere_boot_cumulative_50_2[[i]]$b2[, stat_of_interest]
  observed_plot_mat[, i] <-
    stats_by_sestiere_cumulative_50_2[[i]][[stat_of_interest]]
}

time_series <- 1:nrow(filled_modified)
series_names <- unique(sestiere_df_modified$Sestiere)

# Ensure matrices are data.frames
lower_df <- as.data.frame(b1_plot_mat)
upper_df <- as.data.frame(b2_plot_mat)
observed_df <- as.data.frame(observed_plot_mat)

# Add time column
lower_df$time <- upper_df$time <- observed_df$time <- time_series

# Pivot longer
lower_long <-
  pivot_longer(lower_df,-time, names_to = "series", values_to = "lower")
upper_long <-
  pivot_longer(upper_df,-time, names_to = "series", values_to = "upper")
obs_long   <-
  pivot_longer(observed_df,-time, names_to = "series", values_to = "observed")

# Combine all into one long dataframe
plot_df <- lower_long %>%
  left_join(upper_long, by = c("time", "series")) %>%
  left_join(obs_long, by = c("time", "series")) %>%
  mutate(series = factor(series, levels = paste0("V", 1:6), labels = series_names))

# Calculate the maximum possible entropy for each sestiere
maxima_vec <- rep(0, length(unique(sestiere_df_modified$Sestiere)))
for (s in unique(sestiere_df_modified$Sestiere)) {
  maxima_vec[match(s, unique(sestiere_df_modified$Sestiere))] <-
    log(length(sestiere_df_modified$Parish[sestiere_df_modified$Sestiere ==
                                             s]))
}

# Add these maxima values as a column in the plot dataframe
maxima_column <- rep(maxima_vec, nrow(filled_modified))
plot_df$maxima <- maxima_column

ggplot(plot_df, aes(
  x = rep(filled$Date, each = 6),
  group = series,
  color = series,
  fill = series
)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              alpha = 0.4,
              color = NA) +
  geom_line(aes(y = lower), linetype = "dashed", linewidth = 0.25) + #use dashed line to indicate maximum since color is already being used
  geom_line(aes(y = upper), linetype = "dashed", linewidth = 0.25) +
  geom_line(aes(y = observed), linewidth = 0.5) +
  geom_line(aes(y = maxima), linetype = "dashed", linewidth = 0.75) +
  scale_color_manual(values = c("green", "red", "blue", "yellow", "orange", "pink")) +
  scale_fill_manual(values = c("green", "red", "blue", "yellow", "orange", "pink")) +
  coord_cartesian(ylim = c(0.5, 2.5)) +
  scale_y_continuous(
    name = "Entropy (nats)",
    # Create second axis for effective number
    sec.axis = sec_axis( 
      trans =  ~ .,
      name = "Effective Number",
      labels = function(x)
        scales::comma(exp(x), accuracy = 0.1)
    )
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white")
  ) +
  labs(
    title = "Quantile Function of Cumulative Mortality by Sestiere",
    #title = names(stats_of_whole_cumulative_50_2)[stat_of_interest],
    x = "Date",
    y = "Entropy (nats)",
    color = "Sestiere",
    fill = "Sestiere"
  )

# Save figure

#ggsave("CumulativeQuantile_KL_Sestiere_Final.png",width=2100,height=1500,units="px") #needs to be .jpg
#ggsave("CumulativeQuantile_KL_Sestiere_Final.jpg",width=2100,height=1500,units="px")
#ggsave("~/VenetianPlague_Entropy/Figures/Figure4.png",width=2100,height=1500,units="px")
#ggsave("~/VenetianPlague_Entropy/Figures/Figure4.jpg",width=2100,height=1500,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure4.png",width=2100,height=1500,units="px")
# ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure4.jpg",width=2100,height=1500,units="px")
```

## Figure 1

The goal with Figure 1 is to illustrate how the distribution of disease evolves over the course of the epidemic. Given cumulative mortality data, we represent the distribution of disease in 3 ways: a choropleth map, a quantile function, and a PDF. We then isolate 3 time points corresponding to the start, peak, and end of the epidemic. Finally, we graph these representations at every time point, producing a 3x3 grid.

### Choropleth Maps

The goal of this section is to generate both the choropleth maps as described above for Figure 1 as well as a GIF that cycles through the choropleth map at the end of every month (Supplementary Video S1).

The following code is a modification of G. Lazzari and G. Colavizza's notebook titled "Venice deaths on Maps" published in their 2020 paper. Because the code comes from an older version of R that uses some packages that are no longer supported, the notebook must be run using a separate renv from the rest of this project. As such, we provide the code here for reference purposes but do not run it.

```{r choropleth renv initialization, eval = FALSE}
renv::init(repos="https://packagemanager.posit.co/cran/2020-09-25")
# install.packages("rgdal")
# install.packages("maptools")
# install.packages("tmap")
# install.packages("oce")
```

```{r choropleth load packages, eval = FALSE}
# clean env. vars
rm(list=ls()) 

# packages for maps
library(maptools) # gpclibPermit()
library(tmap)
library(rgdal)
library(oce)
library("RColorBrewer")

# basic packages
library("ggplot2")
library(reshape2)
library(dplyr)
library(lubridate)
```

#### Load shap file for map skeleton

```{r choropleth load shap file, eval = FALSE}
shap = readOGR("~/VenetianPlague_Entropy/1729_PAR_CONF.shp")

# 'id' does not seem to be useful --> remove it
shap@data <- shap@data['NAME']
```

#### Original deaths on map

##### Load clean dataset

```{r choropleth load clean dataset, eval = FALSE}
original_deaths <- read.csv('~/VenetianPlague_Entropy/ALL_utf.csv')
# mind that NA have to be set to 0, ...
# otherwise the aggregrated sum will be reduced to NA!, but...
original_deaths[is.na(original_deaths)] <- 0 # cleaning NAs >> we'll set them back later to NA (***)

# fix dates
original_deaths$Date <- as.POSIXct(original_deaths$Date,format = '%m/%d/%Y')
my_shift <- as.POSIXct('1629-01-01')
original_deaths$Date <- original_deaths$Date - difftime(original_deaths$Date[1],my_shift )
```

##### aggregate over months

(and set back NAs)

```{r choropleth aggregate over months, eval = FALSE}
# melt parishes
melted_df  =  melt(original_deaths,'Date',variable.name = 'parish',value.name = 'deaths')

# group over months, sum ...
# credits to: https://ro-che.info/articles/2017-02-22-group_by_month_r 
agg_melt_df = melted_df %>% group_by(month=floor_date(Date, "month"),parish=parish) %>% 
      summarise(deaths = sum(deaths))

jan_melted_df <- melted_df[melted_df$Date <= as.Date("1629-01-07",format="%Y-%m-%d"),]
jan_agg_melt_df = jan_melted_df %>% group_by(month=floor_date(Date,"month"),parish=parish) %>% summarise(deaths=sum(deaths)) #substitute Jan with up to Jan 7

agg_melt_df[1:70,] <- jan_agg_melt_df

# ...and un-melt again
agg_df = as.data.frame(t(dcast(agg_melt_df,month~parish,drop = FALSE)))

#...and remove useless row
agg_df = agg_df[-1,]

# save parishes names as they will be canceledby next step
parish_names = rownames(agg_df)

# trasform all data to numeric 
agg_df = as.data.frame(sapply(agg_df,function(x){as.numeric(as.character(x))}))

#  (***) set parish with no deaths on the whole t-w to NAs!
agg_df[rowSums(agg_df) == 0,] = NA

# create name col
agg_df$NAME = parish_names

# remove vars of intermediate steps
remove(melted_df,agg_melt_df,parish_names,jan_melted_df,jan_agg_melt_df)
```

##### sum over whole time window

(don't run this section if you are not interested)

create cols for sums and names and drop the rest (and set back NAs)

```{r choropleth sum over whole time window, eval = FALSE}
original_summed <- as.data.frame(colSums(original_deaths[,2:ncol(original_deaths)]))
colnames(original_summed) <- 'Total_deaths'
original_summed$NAME <- rownames(original_summed)
remove(original_deaths)

# remember we had to set NAs to 0 to avoid missing some data). Indeed now we have lost all NAs:
sum(is.na(original_summed))
# so we set it back to NA:
original_summed[original_summed==0] = NA
sum(is.na(original_summed))
```

##### merge deaths data into shape object

fix col names to match names in shapes:

```{r choropleth merge deaths data into shape object, eval = FALSE}
#replace spaces with dots 
shap@data$NAME <- sapply(shap@data$NAME, function(x) gsub('\\s','.',x))
```

fix col names to match names in dataset summed >> over all time-window <<

```{r choropleth fix col names all-time, eval = FALSE}
# the only problematic one is S..Giacomo.dell.Orio --> fixed it: 
original_summed$NAME[original_summed$NAME=='S..Giacomo.dell.Orio'] <- 'S..Giacomo.dell\'Orio'

# check (in)compatibility of names 
sum(!(original_summed$NAME %in% shap@data$NAME)) # -- not bad ;), 0 mismatch!

### view mismatches ### 
View(shap@data[!(shap@data$NAME %in% original_summed$NAME),]) # 15 parishes 
View(original_summed[!(original_summed$NAME %in% shap@data$NAME),])
```

fix col names to match names in dataset aggregated >> over month <<

```{r choropleth fix col names month, eval = FALSE}
# the only problematic one is S..Giacomo.dell.Orio --> fixed it: 
agg_df$NAME[agg_df$NAME=='S..Giacomo.dell.Orio'] <- 'S..Giacomo.dell\'Orio'
## check (in)compatibility of names 
sum(!(agg_df$NAME %in% shap@data$NAME)) # -- not bad ;), 0 mismatch!

### view mismatches ### 
View(shap@data[!(shap@data$NAME %in% agg_df$NAME),]) # 15 parishes have no deaths counts -- it's fine
View(agg_df[!(agg_df$NAME %in% shap@data$NAME),1:2]) 
```

merge data summed over --> all time window <--

```{r choropleth merge data all-time, eval = FALSE}
# keep track of original sorting -- before merging!!
shap@data$id = rownames(shap@data)

# actual merging
shap@data = merge(shap@data,original_summed,
                  by = 'NAME',all.x = TRUE,all.y = TRUE,sort = FALSE)

# set id as numeric 
shap@data$id = as.numeric(as.character(shap@data$id))
# order df using previous id 
shap@data = shap@data[order(shap@data$id),]
```

merge data aggregated --> over months <--

```{r choropleth merge data months, eval = FALSE}

# keep track of original sorting -- before merging!!
#shap@data$id = rownames(shap@data) **Original Code**
shap@data$id = shap@data$id+1

# actual merging
shap@data = merge(shap@data,agg_df,
                  by = 'NAME',all.x = TRUE,sort = FALSE)

# set id as numeric 
shap@data$id = as.numeric(as.character(shap@data$id))
# order df using previous id 
shap@data = shap@data[order(shap@data$id),]

# set deaths data as numeric
shap@data[,3:ncol(shap@data)] = sapply(shap@data[,3:ncol(shap@data)],as.numeric)

# fix col names
months_names <- c(
    "Jan 1629","Feb 1629","Mar 1629","Apr 1629","May 1629","Jun 1629","Jul 1629", "Aug 1629","Sep 1629","Oct 1629", "Nov 1629","Dec 1629", "Jan 1630","Feb 1630","Mar 1630","Apr 1630","May 1630","Jun 1630","Jul 1630", "Aug 1630","Sep 1630","Oct 1630", "Nov 1630","Dec 1630", "Jan 1631","Feb 1631","Mar 1631","Apr 1631","May 1631","Jun 1631","Jul 1631", "Aug 1631","Sep 1631","Oct 1631", "Nov 1631","Dec 1631"
)
# months_names <- c(
#     "Cumulative Mortality up to Jan 1629","Cumulative Mortality up to Feb 1629","Cumulative Mortality up to Mar 1629","Cumulative Mortality up to Apr 1629","Cumulative Mortality up to May 1629","Cumulative Mortality up to Jun 1629","Cumulative Mortality up to Jul 1629", "Cumulative Mortality up to Aug 1629","Cumulative Mortality up to Sep 1629","Cumulative Mortality up to Oct 1629", "Cumulative Mortality up to Nov 1629","Cumulative Mortality up to Dec 1629", "Cumulative Mortality up to Jan 1630","Cumulative Mortality up to Feb 1630","Cumulative Mortality up to Mar 1630","Cumulative Mortality up to Apr 1630","Cumulative Mortality up to May 1630","Cumulative Mortality up to Jun 1630","Cumulative Mortality up to Jul 1630", "Cumulative Mortality up to Aug 1630","Cumulative Mortality up to Sep 1630","Cumulative Mortality up to Oct 1630", "Cumulative Mortality up to Nov 1630","Cumulative Mortality up to Dec 1630", "Cumulative Mortality up to Jan 1631","Cumulative Mortality up to Feb 1631","Cumulative Mortality up to Mar 1631","Cumulative Mortality up to Apr 1631","Cumulative Mortality up to May 1631","Cumulative Mortality up to Jun 1631","Cumulative Mortality up to Jul 1631", "Cumulative Mortality up to Aug 1631","Cumulative Mortality up to Sep 1631","Cumulative Mortality up to Oct 1631", "Cumulative Mortality up to Nov 1631","Cumulative Mortality up to Dec 1631"
# )

colnames(shap@data) <- c(c('NAME','id',"Total_deaths"),months_names)


View(shap@data)
```

##### Plot cases on map

(doc for space analysis/plot in R: <https://cran.r-project.org/doc/contrib/intro-spatial-rl.pdf> )

in order to simplify the plot, we keep only labels for parishes with: N deaths > cut-off

```{r choropleth plot cases, eval = FALSE}
# names_cutoff = 

# for(i in 1:length(original_summed$Total_deaths)){
#   if(is.na(original_summed$Total_deaths[i])){
#     original_summed$Total_deaths[i] <- 0
#   }
# }
# original_summed$Total_deaths
# rownames(original_summed) <- gsub("S..Giacomo.dell.Orio","S..Giacomo.dell\'Orio",rownames(original_summed))

my_name = function(x,cutoff=1000){
    # if(!x %in% rownames(original_summed)){
    #   print(x)
    #   return(NA)
    # }
    # if ((original_summed$Total_deaths[rownames(original_summed)==x] > cutoff) & 
    #     !is.na(original_summed$Total_deaths[rownames(original_summed)==x])){
    #     return(x)
    # }
    if ((shap@data$Total_deaths[shap@data$NAME==x] > cutoff) & 
        !is.na(shap@data$Total_deaths[shap@data$NAME==x])){
        return(x)
    }
    else{
        return(NA)
    }
    
}

shap@data$names_cutoff = sapply(shap@data$NAME,my_name)

# shap@data <- shap@data[-c(71:85),]
#shap@data <- shap@data[order(shap@data$NAME),] #needs to be reordered somehow
invperm <- read.csv("~/LegacyVeniceParishMaps/invperm.csv")
invperm <- invperm$x
#shap@data <- shap@data[shap@plotOrder,]
#shap@data <- shap@data[invperm,]#seems to have no effect while the other permutations of the elements did?
View(shap@data)

```

Import population data and convert to mortality data

```{r choropleth import population data, eval = FALSE}
population_pre <- read.csv("~/VenetianPlague_Entropy/ParishPopulationsPre.csv")
population_post <- read.csv("~/VenetianPlague_Entropy/ParishPopulationsPost.csv")

colnames(population_pre) <- c("Parish", "Population_1509", "Population_1551", "Population_1581", "Population_1586")

population_pre$Parish <- gsub("[[:space:]]",".",population_pre$Parish)
population_post$Parish <- gsub("[[:space:]]",".",population_post$Parish)

non_NAs <- c()
for(i in 1:length(shap@data$NAME)){
  if(length(which(!is.na(shap@data[i,4:ncol(shap@data)]))) != 0){
    non_NAs <- c(non_NAs,i)
  }
}

for(i in 1:length(shap@data$NAME[non_NAs])){
  if(!shap@data$NAME[non_NAs][i] %in% population_pre$Parish){
    print(shap@data$NAME[non_NAs][i])
  }
}

population_pre$Parish <- gsub("S..Giacomo.dell.Orio","S..Giacomo.dell'Orio",population_pre$Parish)

for(i in 1:length(shap@data$NAME[non_NAs])){
  if(!shap@data$NAME[non_NAs][i] %in% population_post$Parish){
    print(shap@data$NAME[non_NAs][i])
  }
}

for(j in 1:ncol(population_pre)){
  if(j == 1){
    next
  }
  for(i in 1:nrow(population_pre)){
    population_pre[i,j] <- format(as.numeric(population_pre[i,j]),nsmall=3)
    if(nchar(population_pre[i,j]) < 7){
      population_pre[i,j] <- as.numeric(population_pre[i,j])*1000
    } else {
      population_pre[i,j] <- as.numeric(population_pre[i,j])
    }
  }
  population_pre[,j] <- as.numeric(population_pre[,j])
}

interpolated_1629 <- as.data.frame(matrix(nrow=length(population_pre$Population_1586),ncol=2))
colnames(interpolated_1629) <- c("Parish","Population")
interpolated_1629$Parish <- population_pre$Parish
interpolated_1629$Population <- (20.45589-(1-0.8990027)*population_pre$Population_1586)*43/110 + population_pre$Population_1586

for(i in 1:nrow(shap@data)){
  if(shap@data$NAME[i] %in% interpolated_1629$Parish){
    # print(shap@data$NAME[i])
    # print(match(shap@data$NAME[i],interpolated_1629$Parish))
    shap@data[i,3:(ncol(shap@data)-1)] <- shap@data[i,3:(ncol(shap@data)-1)]/interpolated_1629$Population[match(shap@data$NAME[i],interpolated_1629$Parish)]
  }
}
```

Convert to cumulative mortality data

```{r choropleth convert to cumulative, eval = FALSE}
for(i in 1:nrow(shap@data)){ #comment this part out if you want instantaneous
  shap@data[i,4:(ncol(shap@data)-1)] <- cumsum(as.numeric(shap@data[i,4:(ncol(shap@data)-1)]))
}
```

map cases >>> by month <<<

```{r choropleth map cases monthly, eval = FALSE}
library(sf)
pdf('map_monthly_cases.pdf',height = 6, width = 12)
#require(sf)

#shap@data <- shap@data[,which(!is.na(colnames(shap@data)))]
#shap@data$Total_deaths <- c(original_summed$Total_deaths[match(rownames(original_summed),shap@data$NAME)],rep(0,15))

sf_use_s2(FALSE) #does this screw something up
# (parishes names makes plot unreadable)

#monthlyshannon <- read.csv("~/LegacyVeniceParishMaps/monthlyshannon.csv")
#monthlyshannon <- read.csv("~/LegacyVeniceParishMaps/monthlyshannonvec2.csv")
monthlyshannon <- read.csv("~/LegacyVeniceParishMaps/monthlyshannonvec3.csv")
monthlyshannon <- monthlyshannon$x

# for(i in 1:length(colnames(shap@data)[20:(ncol(shap@data)-1)])){
#   colnames(shap@data)[20:(ncol(shap@data)-1)][i] <- paste(colnames(shap@data)[20:(ncol(shap@data)-1)][i],round(monthlyshannon[i],2)) 
# }
full_months_names <- c(
    "January 1629","February 1629","March 1629","April 1629","May 1629","June 1629","July 1629", "August 1629","September 1629","October 1629", "November 1629","December 1629", "January 1630","February 1630","March 1630","April 1630","May 1630","June 1630","July 1630", "August 1630","September 1630","October 1630", "November 1630","December 1630", "January 1631","February 1631","March 1631","April 1631","May 1631","June 1631","July 1631", "August 1631","September 1631","October 1631", "November 1631","December 1631"
)

shap@data[1,] <- c("S..Marco",1,rep(as.numeric(NA),38)) 
shap@data[21,] <- c("S..Polo",21,rep(as.numeric(NA),38)) 
shap@data[29,] <- c("S..Stae",29,rep(as.numeric(NA),38))

for(i in 2:39){
  shap@data[,i] <- as.numeric(shap@data[,i])
}

pl <- list()
for(i in (4:(ncol(shap@data)-1))){
  # pl <- c(pl,list(qtm(shap,fill = colnames(shap@data)[i],fill.n=8,fill.breaks=seq(0,0.7,length.out=8))+tm_legend(legend.position=c("right","top"),scale=0.8)))
  pl <- c(pl,list(qtm(shap,fill = colnames(shap@data)[i],fill.n=8,fill.breaks=seq(0,0.7,length.out=8))+tm_legend(legend.position=c("right","top"),scale=0.8)+tm_layout(frame=FALSE))) #,main.title=paste("Distribution of Cumulative Mortality up to",full_months_names[i-3]),main.title.size=3))) #"with Instantaneous Entropy of",round(monthlyshannon[i-3],2))
}

# plot_list3 <- list(plot1=pl[[18]],plot2=pl[[24]],plot3=pl[[34]])
# save(plot_list3,file="plot_list3.RData")
# print(plot_list3)
# plot_list4 <- list(plot1=pl[[6]],plot2=pl[[24]],plot3=pl[[34]])
# save(plot_list4,file="plot_list4.RData")
# print(plot_list4)
# plot_list5 <- list(plot1=pl[[3]],plot2=pl[[22]],plot3=pl[[36]])
# save(plot_list5,file="plot_list5.RData")
# plot_list6 <- list(plot1=pl[[2]],plot2=pl[[22]],plot3=pl[[36]])
# print(plot_list6)
#save(plot_list6,file="plot_list6.RData")
```
Save the plots for Figure 1 (would normally be uncommented)
```{r choropleth save plot, eval = FALSE}
# CURRENT 7_8_26 (START OF UNCOMMENT SECTION)
#plot_list7 <- list(plot1=pl[[1]],plot2=pl[[22]],plot3=pl[[36]])
#print(plot_list7)
#save(plot_list7,file="plot_list7.RData")

# tmap_save(tmap_arrange(pl[[18]],pl[[24]],pl[[34]],ncol=3),filename="gridof3.png",width=2560,height=900,units="px")
# END OF CURRENT 7_8_26 (END OF UNCOMMENT SECTION)
```
Create the GIF (Supplentary Video S1).
```{r choropleth create GIF, eval = FALSE}
#current.mode <- tmap_mode("plot")
#tmap_save(tmap_arrange(pl[[9]],pl[[18]],pl[[27]],pl[[36]]),filename="gridmap_final.png",width=2560,height=1600,units="px")
#tmap_mode(current.mode)


#tmap_animation(pl,width=1920,height=1080,filename="incidencemap_final.gif")

qtm(shap,fill = colnames(shap@data)[20:(ncol(shap@data)-1)],fill.n=8,fill.breaks=seq(0,0.7,length.out=8))

#names_cutoff is wrong because the cutoff is based on total cases not incidence rate
#Also change variable to say Cumulative_incidence_rate not Total_deaths

dev.off()

### testing on single month 
# qtm(shap, text = 'NAME', text.size = 0.5, fill = c('january 1631'))
```

map with labels >>> total cases <<<

```{r choropleth map with labels, eval = FALSE}
pdf('map_total_cases_with_labels.pdf',height = 6, width = 12)

tm_shape(shap) + tm_fill('Total_deaths') + tm_text('NAME', size = 0.5 , col = 'black') + tm_layout(bg.color="lightblue") 

## saving
dev.off()
```

here we fix the text going over boundary -- in the two-row name (fttb, we change only the only problematic one)

```{r choropleth fix text, eval = FALSE}
shap@data$names_cutoff[shap@data$NAME=='S..Pietro.di.Castello'] = 'S..Pietro\ndi.Castello'

View(shap@data[shap@data$NAME=='S..Pietro.di.Castello',])
```

map with labels >>> total cases <<< -- version with less (cutoff) labels

```{r choropleth map reduced labels, eval = FALSE}
pdf('map_total_cases_with_labels_cutoff.pdf',height = 6, width = 12)
# qtm(shap,'Total_deaths',text = 'names_cutoff', text.size = 0.8,text.col = 'black')

tm_shape(shap) + tm_fill('Total_deaths') + tm_text('names_cutoff', size = 0.8 , col = 'black') + tm_layout(bg.color="lightblue")


## saving
dev.off()
```

map total cases without labels

```{r choropleth total cases no labels, eval = FALSE}
# pdf('map_total_cases_no_labels.pdf',height = 6, width = 12)
qtm(shap,'Total_deaths')

#TWO THINGS TO FIX: 1. the names are mixed up 2. the colors are in the wrong place

## saving
dev.off()
```

### PDF and Quantile Graphs

In this section, we will generate the PDF and Quantile Graphs included in Figure 1. For the quantile function and PDF graphs, we overlay the corresponding maximum entropy distribution in red.

First, we define our functions.

```{r PDF_and_Quantile_graphs, eval = FALSE}
# Find the indices in filled$Date of the last day of every month
end_of_month_indices <- c(which(day(filled$Date) == 1)[-1] - 1, nrow(filled_modified))

# Create a new negent function that provides debugging information when NAs are present
negent_new <- function (x, K = ceiling(log2(length(x))) + 1){ # K = number of bins
  p <- table(cut(x, breaks = K))/length(x)
  sigma2 <- var(x)
  #p <- p[(p > 0)]
  bins <- seq(min(x), max(x), length.out = K + 1)
  bins2 <- rep(0, K)
  for(i in 1:K){
    bins2[i] <- (bins[i] + bins[i+1]) / 2
  }
  y <- dnorm(bins2, mean = mean(x), sd = sd(x))
  y <- y / sum(y)
  if(length(which(is.na(y))) == K){
    y <- rep(0, K)
  }
  if(length(as.numeric(p)) != length(y)){
    print("Something went wrong")
    print(paste("p", length(as.numeric(p))))
    print(as.numeric(p))
    print(paste("y", length(y)))
    print(y)
  } else if(anyNA(as.numeric(p))){
    print(paste("p had an NA", as.numeric(p)))
  } else if(anyNA(y)){
    print(paste("y had an NA", y))
  }
  return(KL_new(as.numeric(p), y))
  #return(log(2 * pi * exp(1)*sigma2)*1/2 + sum(p * log(p)))
}

# Create a modified PDF approximation function that adds a small epsilon to avoid NAs
PDF_approx_new <- function (x, K = ceiling(log2(length(x))) + 1, epsilon = 1e-05){
  p <- table(cut(x+epsilon, breaks = K))/length(x)
  #p <- p[(p > 0)]
  return(p)
  #return(log(2 * pi * exp(1)*sigma2)*1/2 + sum(p * log(p)))
}
```

While we use histogram approximation of the PDF with bin number determined by Sturges' rule in the analysis, for the purposes of illustration we use an Epanechnikov kernel with 512 bins in this figure.

```{r PDF_epanechnikov, eval = FALSE}
# Define the PDF approximation function using an Epanechnikov kernel of 512 bins
PDF_approx_epanechnikov <- function (cv){
  #kern <- density(cv,n=512,from=0,to=max(cv),kernel="epanechnikov",bw="bcv")$y
  kern <- density(cv, n = 512, from = min(cv), to = max(cv), bw = "nrd0")$y
  kern <- kern / sum(kern)
  return(kern)
}

#----
# Documentation: normal_approx
# Usage: normal_approx(x, K)
# @description: Takes in a vector of mortality at a given time point and generates a histogram approximation of a Gaussian distribution with identical mean and variance to the data

# Arguments/Options:
# @param x: a vector containing the mortality rates of each parish at a specific point in time
# @param K: the number of bins to use (DEFAULT: ceiling(log2(length(x))) + 1), Sturges' Rule)

# @return: a vector of length K containing the densities in each bin of the normal distribution as described above
#----
normal_approx <- function (x, K = ceiling(log2(length(x))) + 1){
  bins <- seq(min(x), max(x), length.out = K + 1)
  bins2 <- rep(0, K)
  for(i in 1:K){
    bins2[i] <- (bins[i] + bins[i + 1]) / 2
  }
  y <- dnorm(bins2, mean = mean(x), sd = sd(x))
  y <- y / sum(y)
  return(y)
  #return(log(2 * pi * exp(1)*sigma2)*1/2 + sum(p * log(p)))
}

#months_vec <- c(12,24,34)
# Find the indices for 1) January 7, 1629, 2) October 31, 1630, and 3) December 31, 1631
days_vec <- c(7, end_of_month_indices[22], end_of_month_indices[36])

#par(mfrow=c(1,2)) #create matrix
new_negent <- rep(0, length(days_vec))
old_negent <- rep(0, length(days_vec))
shannon_normal <- rep(0, length(days_vec))

quantile_graph_list <- vector(mode = "list", length = 3)
PDF_graph_list <- vector(mode = "list", length = 3)
grobbed_tmap_list <- vector(mode = "list", length = 3)
title_list <- vector(mode = "list", length = 3)

# Choropleth plot
load(file = "~/Downloads/plot_list7.RData")

# for(i in 1:length(end_of_month_indices)){
#   day <- end_of_month_indices[i]
#   case_vec <- as.numeric(colSums(filled_modified[1:day,-1]))/(population_specified_modified$Population*1000)
#   var_mean <- sd(case_vec)/mean(case_vec)
#   plot(PDF_approx_epanechnikov(case_vec),main=paste(filled$Date[day],round(stats_of_whole_cumulative_50_2$negentPDF[day],4)))
# }

# Initialize matrix to calculate coefficient of variation and 95% CI using 999 resamplings
coeff_var_mat <- matrix(NA, nrow = 999, ncol = 3)

for(i in 1:999) {
  indices <- sample(1:ncol(filled_modified[,-1]), replace = TRUE)
  booted_data <- filled_modified[,-1][, indices]
  pop_spec <- population_specified_modified[indices, ]
  for (j in days_vec) {
    day <- j
    case_vec <-
      as.numeric(colSums(booted_data[1:day, ])) / (pop_spec$Population * 1000)
    var_mean <- sd(case_vec) / mean(case_vec)
    coeff_var_mat[i, match(j, days_vec)] <- var_mean
  }
}
sort(coeff_var_mat[,1], decreasing = FALSE)[25] #January 7, 1629 low
sort(coeff_var_mat[,1], decreasing = FALSE)[975] #January 7, 1629 high
sort(coeff_var_mat[,2], decreasing = FALSE)[25] #December 31, 1631 low
sort(coeff_var_mat[,2], decreasing = FALSE)[975] #December 31, 1631 high
sort(coeff_var_mat[,3], decreasing = FALSE)[25] #October 31, 1630 low
sort(coeff_var_mat[,3], decreasing = FALSE)[975] #October 31, 1630 high

for(i in days_vec){
  day <- i
  case_vec <- as.numeric(colSums(filled_modified[1:day,-1])) / (population_specified_modified$Population * 1000)
  var_mean <- sd(case_vec) / mean(case_vec)
  print(var_mean)
  case_vec_norm <- case_vec / sum(case_vec)
  q_func <- sort(case_vec_norm)
  pdf_vec <- as.numeric(PDF_approx_new(case_vec))
  normal <- normal_approx(case_vec)
  shannon_normal[match(i, days_vec)] <- -1 * sum(normal[normal != 0] * log(normal[normal != 0]))
  
  # plot(pdf_vec,main=paste(filled$Date[day],round(var_mean,3))) # PDF
  # lines(normal_approx(case_vec),col="red")
  # barplot(q_func) #quantile
  # plot(PDF_approx_epanechnikov(case_vec))
  
  new_negent[match(i, days_vec)] <- negent_new(case_vec)
  old_negent[match(i, days_vec)] <- negent(case_vec)
  max_ent_uniform <- rep(1/length(case_vec_norm), length(case_vec_norm))
  q_func_df <- as.data.frame(cbind(q_func,max_ent_uniform))
  
  # Generate quantile function graphs 
  
  # Increase y limits for graph from January 7, 1629 since the small number of total cases leads to large proportions in individual parishes
  if(match(i, days_vec) == 1) {
    quantile_graph <-
      ggplot(data = q_func_df, aes(y = q_func, x = seq(1, ncol(filled_modified[, -1])))) +
      geom_bar(stat = "identity") +
      geom_bar(
        aes(y = max_ent_uniform),
        stat = "identity",
        fill = "red",
        alpha = 0.25
      ) +
      labs(y = "Normalized Mortality", x = "Parishes Ranked by Mortality") +
      coord_cartesian(ylim = c(0, 0.15)) +
      theme_bw() +
      theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        axis.title = element_text(size = 10)
      )
    # Only title the first graph to avoid redundancy
    quantile_graph <- quantile_graph + ggtitle("Quantile Function")
  } else {
    quantile_graph <-
      ggplot(data = q_func_df, aes(y = q_func, x = seq(1, ncol(filled_modified[, -1])))) +
      geom_bar(stat = "identity") +
      geom_bar(
        aes(y = max_ent_uniform),
        stat = "identity",
        fill = "red",
        alpha = 0.25
      ) +
      labs(y = "Normalized Mortality", x = "Parishes Ranked by Mortality") +
      coord_cartesian(ylim = c(0, 0.04)) +
      theme_bw() +
      theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        axis.title = element_text(size = 10)
      )
  }
  
  # Generate PDF graphs
  
  K <- ceiling(log2(length(case_vec))) + 1
  bins <- seq(min(case_vec), max(case_vec), length.out = K + 1)
  bins2 <- rep(0, K)
  for(j in 1:K){
    bins2[j] <- (bins[j] + bins[j+1]) / 2
  }
  bins3 <- seq(from = min(case_vec), to = max(case_vec), length.out = 512)
  pdf_vec_2 <- PDF_approx_epanechnikov(case_vec)
  max_ent_normal <- dnorm(bins3, mean = mean(case_vec), sd = sd(case_vec))
  max_ent_normal <- max_ent_normal / sum(max_ent_normal)
  pdf_df <- data.frame(x = bins3, y = pdf_vec_2, y2 = max_ent_normal)
  
  # Increase y limits for the graph on January 6, 1629 for the same reason as before
  if(match(i, days_vec) == 1) {
    PDF_graph <- ggplot(data = pdf_df, aes(x = x, y = y)) +
      geom_density(
        stat = "identity",
        fill = "darkgrey",
        alpha = 0.5,
        lwd = 0.25
      ) +
      geom_density(
        aes(y = y2),
        stat = "identity",
        fill = "red",
        alpha = 0.25,
        lwd = 0.25
      ) +
      labs(y = "Normalized Parish Count", x = "Cumulative Mortality") +
      coord_cartesian(ylim = c(0, 0.01)) +
      theme_bw() +
      theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        axis.title = element_text(size = 10)
      )
    # Only title the first graph to avoid redundancy
    PDF_graph <- PDF_graph + ggtitle("PDF")
  } else {
    PDF_graph <- ggplot(data = pdf_df, aes(x = x, y = y)) +
      geom_density(
        stat = "identity",
        fill = "darkgrey",
        alpha = 0.5,
        lwd = 0.25
      ) +
      geom_density(
        aes(y = y2),
        stat = "identity",
        fill = "red",
        alpha = 0.25,
        lwd = 0.25
      ) +
      labs(y = "Normalized Parish Count", x = "Cumulative Mortality") +
      coord_cartesian(ylim = c(0, 0.006)) +
      theme_bw() +
      theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        axis.title = element_text(size = 10)
      )
  }
  quantile_graph_list[[match(i, days_vec)]] <- quantile_graph
  
  if(match(i,days_vec) == 1){
    tmap_plot <- plot_list7[[match(i, days_vec)]] +
      tm_layout(
        # Add spaces to the title to align the title with the other two
        main.title="    Choropleth Map",
        main.title.size=1.1,
        #legend.text.size = .55,
        outer.margins = c(0, 0, 0, 0),
        inner.margins = c(0, 0, 0, 0),
        frame = FALSE
      )
  } else{
    tmap_plot <- plot_list7[[match(i, days_vec)]] +
      tm_layout(
        main.title = " ",
        main.title.size = 1.1,
        #legend.text.size = .55,
        outer.margins = c(0, 0, 0, 0),
        inner.margins = c(0, 0, 0, 0),
        frame = FALSE
      )
  }
  grobbed_tmap_plot <- tmap_grob(tmap_plot)

  grobbed_tmap_list[[match(i, days_vec)]] <- grobbed_tmap_plot

  PDF_graph_list[[match(i, days_vec)]] <- PDF_graph
  
  PDF_and_quantile_graph <- plot_grid(plotlist = list(PDF_graph,quantile_graph))
  
  #untitled_grid <- plot_grid(plotlist=list(PDF_and_quantile_graph,heatmap_plot))
  
  title <- ggdraw() + 
    draw_label(
      filled$Date[day],
      fontface = 'bold',
      x = 0,
      hjust = 0
    ) +
    theme(
      # add margin on the left of the drawing canvas,
      # so title is aligned with left edge of first plot
      plot.margin = margin(0, 0, 0, 7)
    )
  title_list[[match(i,days_vec)]] <- title
  overall_graph <- plot_grid(
    title, PDF_and_quantile_graph,
    ncol = 1, nrow = 2,
    # rel_heights values control vertical title margins
    rel_heights = c(0.1, 1)
  )
  print(overall_graph)
}
par(mfrow = c(1, 1)) #reset default gridding layout

# Initialize lists to construct the 3x3 grid
master_plot_list <- vector(mode = "list", length = 9)
master_plot_list_1 <- vector(mode = "list", length = 3)
master_plot_list_2 <- vector(mode = "list", length = 3)
master_plot_list_3 <- vector(mode = "list", length = 3)

# First, append all of the choropleth maps
for(k in 1:length(days_vec)){
  master_plot_list[[k]] <- grobbed_tmap_list[[k]]
  master_plot_list_1[[k]] <- grobbed_tmap_list[[k]]
}

# Next, append all of the quantile function figures
for(k in 4:6){
  master_plot_list[[k]] <- quantile_graph_list[[k-3]]
  master_plot_list_2[[k-3]] <- quantile_graph_list[[k-3]]
}

# Finally, append all of the PDF figures
for(k in 7:9){
  master_plot_list[[k]] <- PDF_graph_list[[k-6]]
  master_plot_list_3[[k-6]] <- PDF_graph_list[[k-6]]
}

# Group by row and give the corresponding label to the first graph of every row
plot_grid_1 <- plot_grid(plotlist = master_plot_list_1, ncol = 3, nrow = 1, align = "hv", axis = "tblr", labels = c("a", "", ""), label_size = 12)
plot_grid_2 <- plot_grid(plotlist = master_plot_list_2, ncol = 3, nrow = 1, align = "hv", axis = "tblr", labels = c("b", "", ""), label_size = 12)
plot_grid_3 <- plot_grid(plotlist = master_plot_list_3, ncol = 3, nrow = 1, align = "hv", axis = "tblr", labels = c("c", "", ""), label_size = 12)

# Generate the final plot as a group of the three rows
plot_grid(plot_grid_1, plot_grid_2, plot_grid_3, nrow = 3, align = "hv", axis = "tblr")

plot_grid(plotlist = master_plot_list, ncol = 3, nrow = 3, align = 'hv', axis = 'tblr', labels = c("a", "", "", "b", "", "", "c", "", ""), label_size = 12)

# Save the figure

#ggsave("Gridded_Incidence_Map_Final.png",width=2900,height=1800,units="px")
#ggsave("Gridded_Incidence_Map_Final_10.png", width = 3000, height = 1800, units = "px")
#ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure1v2.png", width = 3000, height = 1800, units = "px")
#ggsave("~/VenetianPlague_Entropy/Figures_Final/Figure1v2.jpg", width = 3000, height = 1800, units = "px")

# plot.ts(new_negent,col="red",ylim=c(0,2))
# lines(old_negent)
```

## Supplementary Figure S7 (Mortality Graphs)

In this section, we generate the mortality graphs seen in Supplementary Figure S7.

```{r mortality graphs, eval = FALSE}
outliers <- c("S.Polo", "S.Stae","S.Marco") #No cases between 9/1/30 and 12/31/30, outbreak time

filled_nodate <- filled[,-1]
outlier_indices <- which(colnames(filled_nodate) %in% outliers)
total_deaths <- colSums(filled_nodate[1:nrow(filled_modified),])

print(colnames(filled_nodate) == population_specified$Parish)

export_df <- as.data.frame(cbind(1:ncol(filled_nodate), total_deaths, population_specified$Population * 1000))
write.csv(export_df, "total_deaths.csv")

print(outlier_indices)

mort_df <- as.data.frame(matrix(NA, ncol = 3, nrow = nrow(filled_modified)))
colnames(mort_df) <- c("Date", "Mort", "CumMort")
mort_df$Date <- as.Date(filled$Date)

for(i in 1:nrow(filled_modified)){
  mort_df[i,2] <- sum(as.numeric(filled_nodate[i,]) / (population_specified$Population * 1000)) / ncol(filled_nodate)
  mort_df[i,3] <- sum(as.numeric(colSums(filled_nodate[1:i,])) / (population_specified$Population * 1000)) / ncol(filled_nodate)
}

regions <- tibble(x1 = as.Date("1630-09-01"), x2 = as.Date("1630-12-31"), y1 = -Inf, y2 = +Inf)

daily_mort_plot <- ggplot(mort_df)+
  geom_rect(data = regions, inherit.aes = FALSE, mapping = aes(xmin = x1, xmax = x2, ymin = y1, ymax = y2),color = "transparent",fill = "blue", alpha = 0.2)+
  geom_line(aes(x = as.Date(Date), y = Mort))+
  labs(x = "Date", y = "Mean Daily Parish Mortality", title = "Daily Mortality")+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.title = element_text(size = 10))

cumulative_mort_plot <- ggplot(mort_df)+
  geom_rect(data = regions, inherit.aes = FALSE, mapping = aes(xmin = x1, xmax = x2, ymin = y1, ymax = y2), color = "transparent", fill = "blue", alpha = 0.2)+
  geom_line(aes(x = as.Date(Date), y = CumMort))+
  labs(x = "Date", y = "Mean Cumulative Parish Mortality", title = "Cumulative Mortality")+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.title = element_text(size = 10))

untitled_grid <- plot_grid(plotlist = list(daily_mort_plot, cumulative_mort_plot),labels = c("a", "b"), label_size = 12)

title <- ggdraw() + 
  draw_label(
    "Mortality Over Time",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    # add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7)
  )
plot_grid(
  title, untitled_grid,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 1)
)
#ggsave("SuppFig7.jpg", width = 2200, height = 1080, units = "px")

sestiere_vec <- c()
for(i in unique(sestiere_df_modified$Sestiere)){
  sestiere_vec <- c(sestiere_vec, length(sestiere_df_modified$Parish[sestiere_df_modified$Sestiere==i]))
}
print(sum(sestiere_vec) / length(sestiere_vec))
print(sd(sestiere_vec))

final_mortality <- as.numeric(colSums(filled_modified[1:nrow(filled_modified),-1]) / (population_specified_modified$Population * 1000))
mean_mortality <- c()
for(i in unique(sestiere_df_modified$Sestiere)){
  sestiere_mortality <- final_mortality[which(population_specified_modified$Parish %in% sestiere_df_modified$Parish[sestiere_df_modified$Sestiere == i])]
  mean_mortality <- c(mean_mortality, sum(sestiere_mortality) / length(sestiere_mortality))
  print(sd(sestiere_mortality))
}
print(sum(mean_mortality) / length(mean_mortality))
print(sd(mean_mortality))
```

## Testing Binning Methods

In this section, we compare the entropy of the PDF with that of the histogram-approximated normal distribution with identical mean and variance to the data. In particular, we make this comparison for three bin selection methods.

### Sturges' Rule

Here, we implement Sturges' Rule to determine the number of bins ($K = \lceil \log_2(n)\rceil + 1$). Since we have a consistent number of parishes ($n = 50$), Sturges' Rule will always result in a 7 bin histogram. 

```{r find PDF max entropy sturges rule}
normal_approx <- function (x, K = ceiling(log2(length(x))) + 1){
  bins <- seq(min(x), max(x), length.out = K + 1)
  bins2 <- rep(0, K)
  for(i in 1:K){
    bins2[i] <- (bins[i] + bins[i + 1]) / 2
  }
  y <- dnorm(bins2, mean = mean(x), sd = sd(x))
  y <- y / sum(y)
  plot(x = bins2, y = y)
  y_2 <- diff(pnorm(bins, mean(x), sd(x)))
  plot(bins2, y_2)
  print(y_2 - y)
  print(mean(x) - sum(bins2 * y))
  print(mean(x) - sum(bins2 * y_2))
  print(sum(y_2))
  return(y)
  #return(log(2 * pi * exp(1)*sigma2)*1/2 + sum(p * log(p)))
}

for(i in 1:10){
  case_vector_2 <- colSums(filled_modified[1:i,-1])/(population_specified_modified$Population * 1000)
  p <- normal_approx(case_vector_2)
}

#shannon_PDF_vec <- rep(0, nrow(filled_modified))
#case_df <- as.data.frame(matrix(data = NA, nrow = nrow(filled_modified), ncol = ncol(filled_modified[,-1])))
#for(i in 1:nrow(filled_modified)){
#  case_vector_2 <- colSums(filled_modified[1:i,-1])/(population_specified_modified$Population * 1000)
#  case_df[i,] <- case_vector_2
#  p <- normal_approx(case_vector_2)
#  max_entropy <- -1 * sum(p*log(p))
#  #kern <- as.numeric(PDF_approx(case_vector_2))
#  kern <- as.numeric(hist(case_vector_2, breaks = seq(min(case_vector_2), max(case_vector_2), length.out = 8), plot = FALSE)$density)
#  kern <- kern/sum(kern)
#  kern_norm <- kern / sum(kern)
#  shannon_PDF <- -1 * sum(kern[kern != 0] * log(kern[kern != 0]))
#  max_entropy_vec[i] <- max_entropy
#  shannon_PDF_vec[i] <- shannon_PDF
#}
```

Now, we plot:
1. the entropy of the resulting PDF (black) against that of the normal distribution with the same mean and variance (red) for every day in the epidemic
2. the negentropy of the PDF, calculated as entropy of the normal distribution - entropy of the PDF

```{r plot PDF vs normal Sturges rule, eval = FALSE}
plot_df <- as.data.frame(cbind(shannon_PDF_vec, max_entropy_vec, filled$Date))
plot1 <- ggplot(data = plot_df, aes(x = filled$Date, y = shannon_PDF_vec)) +
  geom_line() +
  geom_line(aes(x = filled$Date, y = max_entropy_vec), col = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Entropy (nats)", x = "Date", title = "Entropy")

plot2 <- ggplot(data = plot_df, aes(x = filled$Date, y = max_entropy_vec - shannon_PDF_vec)) +
  geom_line() +
  geom_hline(yintercept = 0, color = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Negentropy (nats)", x = "Date", title = "Negentropy")

plot_ls <- list(plot1, plot2)

# Create the grid 
untitled_grid <-
  plot_grid(plotlist = plot_ls,
            labels = c("a", "b"),
            label_size = 12)

# Add the title
title <- ggdraw() +
  draw_label(
    "PDF of Cumulative Mortality Using Sturges' Rule",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(# add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7))

plot_grid(title,
          untitled_grid,
          ncol = 1,
          # rel_heights values control vertical title margins
          rel_heights = c(0.1, 1))
```

### Freedman-Diaconis

Here, we implement the Freedman-Diaconis rule to determine our bin width. The Freedman-Diaconis rule states that $\text{Bin width} = 2 \cdot \frac{\text{IQR}(x)}{\sqrt[3]{x}}$. 

```{r find PDF max entropy using freedman diaconis, eval = FALSE}
freedman_diaconis <- function (x){
  bin_width = 2 * IQR(x) * (length(x))^(-1/3)
  num_bins = ceiling((max(x) - min(x))/bin_width)
  return(num_bins)
}

max_entropy_vec <- rep(NA_real_, nrow(filled_modified))
shannon_PDF_vec <- rep(NA_real_, nrow(filled_modified))

# Exclude the first day since the high prevalence of zeroes results in 
for(i in 2:nrow(filled_modified)){
  case_vector_2 <- colSums(filled_modified[1:i,-1])/(population_specified_modified$Population * 1000)
  num_bins <- as.numeric(freedman_diaconis(case_vector_2))
  p <- normal_approx(case_vector_2, K = num_bins)
  max_entropy <- -1 * sum(p*log(p))
  kern <- as.numeric(hist(case_vector_2, breaks = seq(min(case_vector_2), max(case_vector_2), length.out = num_bins + 1), plot = FALSE)$density)
  kern <- kern/sum(kern)
  #if(i %% 100 == 0){
   # hist(case_vector_2, breaks = num_bins)
  #  print(hist(case_vector_2, breaks = num_bins, plot = FALSE)$density)
  #}
  #kern <- as.numeric(PDF_approx(case_vector_2, K = num_bins))
  kern_norm <- kern / sum(kern)
  shannon_PDF <- -1 * sum(kern[kern != 0] * log(kern[kern != 0]))
  
  max_entropy_vec[i] <- max_entropy
  shannon_PDF_vec[i] <- shannon_PDF
}

```

Now, we plot the entropy of the resulting PDF (black) against that of the normal distribution with the same mean and variance (red) for every day in the epidemic.

```{r plot PDF vs normal Freedman-Diaconis, eval = FALSE}
plot_df<- as.data.frame(cbind(shannon_PDF_vec, max_entropy_vec, filled$Date))
plot1 <- ggplot(data = plot_df, aes(x = filled$Date, y = shannon_PDF_vec)) +
  geom_line() +
  geom_line(aes(x = filled$Date, y = max_entropy_vec), col = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Entropy (nats)", x = "Date", title = "Entropy")

plot2 <- ggplot(data = plot_df, aes(x = filled$Date, y = max_entropy_vec - shannon_PDF_vec)) +
  geom_line() +
  geom_hline(yintercept = 0, color = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Negentropy (nats)", x = "Date", title = "Negentropy")

plot_ls <- list(plot1, plot2)

# Create the grid 
untitled_grid <-
  plot_grid(plotlist = plot_ls,
            labels = c("a", "b"),
            label_size = 12)

# Add the title
title <- ggdraw() +
  draw_label(
    "PDF of Cumulative Mortality Using Freedman-Diaconis Rule",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(# add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7))

plot_grid(title,
          untitled_grid,
          ncol = 1,
          # rel_heights values control vertical title margins
          rel_heights = c(0.1, 1))
```

### Kernel-Density Estimation

Finally, here we use Kernel-Density estimation with 512 bins. Note that because the total number of bins is substantially greater than the previous two methods, the magnitude of entropy will also be larger. 

```{r kde entropy}
kde_entropy <- function(x, n_points = 512) {
  x <- x[is.finite(x)]

  if (length(x) < 2L || length(unique(x)) < 2L) {
    return(c(entropy = NA_real_,
             maximum_entropy = NA_real_,
             negentropy = NA_real_))
  }

  d <- density(x, n = n_points)
  dx <- d$x[2L] - d$x[1L]

  f <- d$y / sum(d$y * dx)
  positive <- f > 0

  h <- -sum(f[positive] * log(f[positive])) * dx

  mu <- sum(d$x * f) * dx
  variance <- sum((d$x - mu)^2 * f) * dx
  h_normal <- 0.5 * log(2 * pi * exp(1) * variance)

  c(
    entropy = h,
    maximum_entropy = h_normal,
    negentropy = h_normal - h
  )
}

max_entropy_vec <- rep(0, nrow(filled_modified))
shannon_PDF_vec <- rep(0, nrow(filled_modified))

for (i in 1:nrow(filled_modified)) {
  case_vector_2 <- colSums(filled_modified[1:i, -1]) / (population_specified_modified$Population * 1000)
  kde_output_vec <- kde_entropy(case_vector_2)
  shannon_PDF_vec[i] <- kde_output_vec[1]
  max_entropy_vec[i] <- kde_output_vec[2]
}
```

Now, we plot the entropy of the resulting PDF (black) against that of the normal distribution with the same mean and variance (red) for every day in the epidemic.

```{r kde entropy plot}
plot_df <- as.data.frame(cbind(shannon_PDF_vec, max_entropy_vec, filled$Date))
plot1 <- ggplot(data = plot_df, aes(x = filled$Date, y = shannon_PDF_vec)) +
  geom_line() +
  geom_line(aes(x = filled$Date, y = max_entropy_vec), col = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Entropy (nats)", x = "Date", title = "Entropy")

plot2 <- ggplot(data = plot_df, aes(x = filled$Date, y = max_entropy_vec - shannon_PDF_vec)) +
  geom_line() +
  geom_line(aes(x = filled$Date, y = rep(0, nrow(filled_modified))), col = "red") +
  scale_x_date(date_breaks = "12 months", date_labels = "%m-%Y") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "white"),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  ) +
  labs(y = "Negentropy (nats)", x = "Date", title = "Negentropy")

plot_ls <- list(plot1, plot2)

# Create the grid 
untitled_grid <-
  plot_grid(plotlist = plot_ls,
            labels = c("a", "b"),
            label_size = 12)

# Add the title
title <- ggdraw() +
  draw_label(
    "PDF of Cumulative Mortality Using KDE",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(# add margin on the left of the drawing canvas,
    # so title is aligned with left edge of first plot
    plot.margin = margin(0, 0, 0, 7))

plot_grid(title,
          untitled_grid,
          ncol = 1,
          # rel_heights values control vertical title margins
          rel_heights = c(0.1, 1))
```

### Binning Method Conclusions

Note that regardless of the binning method used, negentropy starts low, increases until the primary outbreak, then drops and stays low for the rest of the epidemic.

## Dynamic Parish Populations

In this section, we repeat our entropy calculations using dynamic parish populations; specifically, the population of a parish on day $d$ is represented as the population of that parish on January 1, 1629 minus the total deaths in that parish up to but not including day $d$. Ignoring births, the purpose of this analysis is to quantify the extent to which the assumption of fixed parish populations artifactually drives the distribution of cumulative mortality towards uniformity.

Non-dynamic populations (baseline).
```{r non-dynamic parish populations}
instantaneous_mortality_df <- as.data.frame(matrix(NA_real_, nrow = nrow(filled_modified), ncol = ncol(filled_modified) - 1))

shannonPDF_vec <- rep(0, nrow(filled_modified))
shannonQuantile_vec <- rep(0, nrow(filled_modified))
negentPDF_vec <- rep(0, nrow(filled_modified))
uniformKL_Quantile_vec <- rep(0, nrow(filled_modified))

for(i in 1:nrow(filled_modified)){
  instantaneous_cv <- filled_modified[i, -1] / (population_specified_modified$Population * 1000)
  instantaneous_mortality_df[i, ] <- instantaneous_cv
  
  cumulative_cv <- colSums(instantaneous_mortality_df[1:i,])
  cumulative_cv <- sort(cumulative_cv)
    
  # Normalized case vector (cumulative_cv_norm): quantile function
  cumulative_cv_norm <- cumulative_cv / sum(cumulative_cv)
  
  kde_output_vec_2 <- kde_entropy(cumulative_cv)
    
  shannon_PDF <- kde_output_vec_2[1]
  max_entropy_PDF <- kde_output_vec_2[2]
  
  shannon_quantile <- -1 * sum(cumulative_cv_norm[cumulative_cv_norm != 0] * log(cumulative_cv_norm[cumulative_cv_norm != 0]))
    
  negent_PDF <- max_entropy_PDF - shannon_PDF
  
  uniform_KL_quantile <- KL_new(cumulative_cv_norm, rep(1 / length(cumulative_cv_norm), length(cumulative_cv_norm)))

  shannonPDF_vec[i] <- shannon_PDF
  shannonQuantile_vec[i] <- shannon_quantile
  negentPDF_vec[i] <- negent_PDF
  uniformKL_Quantile_vec[i] <- uniform_KL_quantile
}
```

Dynamic populations.
```{r dynamic parish populations}
population_specified_dynamic <- population_specified_modified
instantaneous_dynamic_mortality_df <- as.data.frame(matrix(NA_real_, nrow = nrow(filled_modified), ncol = ncol(filled_modified) - 1))

shannonPDF_dynamic_vec <- rep(0, nrow(filled_modified))
shannonQuantile_dynamic_vec <- rep(0, nrow(filled_modified))
negentPDF_dynamic_vec <- rep(0, nrow(filled_modified))
uniformKL_Quantile_dynamic_vec <- rep(0, nrow(filled_modified))

for(i in 1:nrow(filled_modified)){
  instantaneous_cv <- filled_modified[i, -1] / (population_specified_dynamic$Population * 1000)
  instantaneous_dynamic_mortality_df[i, ] <- instantaneous_cv
  
  cumulative_cv <- colSums(instantaneous_dynamic_mortality_df[1:i,])
  cumulative_cv <- sort(cumulative_cv)
    
  # Normalized case vector (cumulative_cv_norm): quantile function
  cumulative_cv_norm <- cumulative_cv / sum(cumulative_cv)
  
  kde_output_vec_2 <- kde_entropy(cumulative_cv)
    
  shannon_PDF <- kde_output_vec_2[1]
  max_entropy_PDF <- kde_output_vec_2[2]
  
  shannon_quantile <- -1 * sum(cumulative_cv_norm[cumulative_cv_norm != 0] * log(cumulative_cv_norm[cumulative_cv_norm != 0]))
    
  negent_PDF <- max_entropy_PDF - shannon_PDF
  
  uniform_KL_quantile <- KL_new(cumulative_cv_norm, rep(1 / length(cumulative_cv_norm), length(cumulative_cv_norm)))

  shannonPDF_dynamic_vec[i] <- shannon_PDF
  shannonQuantile_dynamic_vec[i] <- shannon_quantile
  negentPDF_dynamic_vec[i] <- negent_PDF
  uniformKL_Quantile_dynamic_vec[i] <- uniform_KL_quantile
  
  population_specified_dynamic$Population <- population_specified_dynamic$Population - (as.numeric(filled_modified[i,-1]) / 1000)
}
```

Plot the results

```{r plot dynamic parish populations}
plot.ts(shannonPDF_vec)
lines(shannonPDF_dynamic_vec, col = "red")

plot.ts(shannonQuantile_vec, ylim = c(3.5, 3.9))
lines(shannonQuantile_dynamic_vec, col = "red")

plot.ts(negentPDF_vec)
lines(negentPDF_dynamic_vec, col = "red")

plot.ts(uniformKL_Quantile_vec, ylim = c(0, 0.4))
lines(uniformKL_Quantile_dynamic_vec, col = "red")
```

The change in quantile function entropy after the outbreak is mitigated. This raises the question of what might happen if we start the analysis on the first day of the outbreak.

Start on the first day of the outbreak.

```{r start on outbreak}
num_zeroes <- rep(0, nrow(filled_modified))
for(i in 1:nrow(filled_modified)){
  num_zeroes[i] <- sum(filled_modified[i,-1] == 0)
}
print(match(min(num_zeroes), num_zeroes))
print(match(as.Date("1630-09-01", format = "%Y-%m-%d"), filled_modified$Date))

new_start_date <- 637
```

Non-dynamic populations (baseline).
```{r start on outbreak non-dynamic parish populations}
instantaneous_mortality_df <- as.data.frame(matrix(NA_real_, nrow = nrow(filled_modified), ncol = ncol(filled_modified) - 1))

shannonPDF_vec <- rep(NA_real_, nrow(filled_modified))
shannonQuantile_vec <- rep(NA_real_, nrow(filled_modified))
negentPDF_vec <- rep(NA_real_, nrow(filled_modified))
uniformKL_Quantile_vec <- rep(NA_real_, nrow(filled_modified))

for(i in new_start_date:nrow(filled_modified)){
  instantaneous_cv <- filled_modified[i, -1] / (population_specified_modified$Population * 1000)
  instantaneous_mortality_df[i, ] <- instantaneous_cv
  
  cumulative_cv <- colSums(instantaneous_mortality_df[new_start_date:i,])
  cumulative_cv <- sort(cumulative_cv)
    
  # Normalized case vector (cumulative_cv_norm): quantile function
  cumulative_cv_norm <- cumulative_cv / sum(cumulative_cv)
  
  kde_output_vec_2 <- kde_entropy(cumulative_cv)
  
  shannon_PDF <- kde_output_vec_2[1]
  max_entropy_PDF <- kde_output_vec_2[2]
  
  shannon_quantile <- -1 * sum(cumulative_cv_norm[cumulative_cv_norm != 0] * log(cumulative_cv_norm[cumulative_cv_norm != 0]))
    
  negent_PDF <- max_entropy_PDF - shannon_PDF
  
  uniform_KL_quantile <- KL_new(cumulative_cv_norm, rep(1 / length(cumulative_cv_norm), length(cumulative_cv_norm)))

  shannonPDF_vec[i] <- shannon_PDF
  shannonQuantile_vec[i] <- shannon_quantile
  negentPDF_vec[i] <- negent_PDF
  uniformKL_Quantile_vec[i] <- uniform_KL_quantile
}
```

Dynamic populations.
```{r start on outbreak dynamic parish populations}
population_specified_dynamic <- population_specified_modified
instantaneous_dynamic_mortality_df <- as.data.frame(matrix(NA_real_, nrow = nrow(filled_modified), ncol = ncol(filled_modified) - 1))

shannonPDF_dynamic_vec <- rep(NA_real_, nrow(filled_modified))
shannonQuantile_dynamic_vec <- rep(NA_real_, nrow(filled_modified))
negentPDF_dynamic_vec <- rep(NA_real_, nrow(filled_modified))
uniformKL_Quantile_dynamic_vec <- rep(NA_real_, nrow(filled_modified))

for(i in new_start_date:nrow(filled_modified)){
  instantaneous_cv <- filled_modified[i, -1] / (population_specified_dynamic$Population * 1000)
  instantaneous_dynamic_mortality_df[i, ] <- instantaneous_cv
  
  cumulative_cv <- colSums(instantaneous_dynamic_mortality_df[new_start_date:i,])
  cumulative_cv <- sort(cumulative_cv)
    
  # Normalized case vector (cumulative_cv_norm): quantile function
  cumulative_cv_norm <- cumulative_cv / sum(cumulative_cv)
  
  kde_output_vec_2 <- kde_entropy(cumulative_cv)
    
  shannon_PDF <- kde_output_vec_2[1]
  max_entropy_PDF <- kde_output_vec_2[2]
  
  shannon_quantile <- -1 * sum(cumulative_cv_norm[cumulative_cv_norm != 0] * log(cumulative_cv_norm[cumulative_cv_norm != 0]))
    
  negent_PDF <- max_entropy_PDF - shannon_PDF
  
  uniform_KL_quantile <- KL_new(cumulative_cv_norm, rep(1 / length(cumulative_cv_norm), length(cumulative_cv_norm)))

  shannonPDF_dynamic_vec[i] <- shannon_PDF
  shannonQuantile_dynamic_vec[i] <- shannon_quantile
  negentPDF_dynamic_vec[i] <- negent_PDF
  uniformKL_Quantile_dynamic_vec[i] <- uniform_KL_quantile
  
  population_specified_dynamic$Population <- population_specified_dynamic$Population - (as.numeric(filled_modified[i,-1]) / 1000)
}
```

Plot the results

```{r start on outbreak plot dynamic parish populations}
plot.ts(shannonPDF_vec, xlim = c(new_start_date, nrow(filled_modified)), ylim = c(-7, 0))
lines(shannonPDF_dynamic_vec, col = "red")

plot.ts(shannonQuantile_vec, xlim = c(new_start_date, nrow(filled_modified)))
lines(shannonQuantile_dynamic_vec, col = "red")

plot.ts(negentPDF_vec, xlim = c(new_start_date, nrow(filled_modified)), ylim = c(0, 0.2))
lines(negentPDF_dynamic_vec, col = "red")
lines(rowSums(filled_modified[,-1])/1000, col = "blue")

plot.ts(uniformKL_Quantile_vec, xlim = c(new_start_date, nrow(filled_modified)), ylim = c(0, 0.4))
lines(uniformKL_Quantile_dynamic_vec, col = "red")
lines(rowSums(filled_modified[,-1])/2000, col = "blue")
```